---
title: "Categorical - ADVANCED"
subtitle: "Hierarchical & Complex Categorical Data"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Advanced Categorical Visualization {.center}

## What We'll Learn - Advanced

::: {.columns}
::: {.column width="50%"}
**Complex Categorical Structures:**

- Treemaps (hierarchical proportions)
- Sunburst charts (circular hierarchies)
- Dendrograms (clustering)
- Marimekko charts (two-way composition)
- Waffle charts (unit visualization)
:::

::: {.column width="50%"}
**Analytical Goals:**

- 🔴 **Part of Whole** - Hierarchical composition
- 🟢 **Ranking** - Within hierarchies
- ⚫ **Correlation** - Category associations
- 🟡 **Distribution** - Nested patterns
:::
:::

## Setup Code

```{r}
#| echo: true
#| message: false

# Load required packages
library(readxl)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(treemapify)  # For treemaps
library(sunburstR)   # For sunburst charts
library(ggmosaic)    # For mosaic plots

# Load dataset
health_data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Create hierarchical data
hierarchical_data <- health_data %>%
  filter(!is.na(Region), !is.na(Gender), !is.na(SmokingStatus)) %>%
  count(Region, Gender, SmokingStatus, name = "count")

# Create theme
theme_custom <- function() {
  theme_minimal() +
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "bottom"
  )
}
```

## Treemap - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show hierarchical part-to-whole relationships

**Goal:** 🔴 Part of Whole + 🟢 Ranking

```{r}
#| echo: true
#| eval: false

# Prepare data for treemap
treemap_data <- health_data %>%
  filter(!is.na(Region), !is.na(TreatmentGroup)) %>%
  count(Region, TreatmentGroup, name = "count") %>%
  mutate(label = paste0(TreatmentGroup, "\n(n=", count, ")"))

# Create treemap
ggplot(treemap_data, aes(area = count, fill = Region, 
                         label = label, subgroup = Region)) +
  geom_treemap() +
  geom_treemap_subgroup_border(color = "white", size = 3) +
  geom_treemap_subgroup_text(
    place = "center",
    grow = TRUE,
    alpha = 0.5,
    color = "black",
    fontface = "bold",
    size = 18
  ) +
  geom_treemap_text(
    color = "white",
    place = "center",
    size = 10,
    grow = FALSE
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Patient Distribution by Region and Treatment",
       subtitle = "Rectangle size proportional to patient count",
       fill = "Region") +
  theme_custom() +
  theme(legend.position = "right")
```
:::

## Treemap - Output

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| warning: false

treemap_data <- health_data %>%
  filter(!is.na(Region), !is.na(TreatmentGroup)) %>%
  count(Region, TreatmentGroup, name = "count") %>%
  mutate(label = paste0(TreatmentGroup, "\n(n=", count, ")"))

ggplot(treemap_data, aes(area = count, fill = Region, 
                         label = label, subgroup = Region)) +
  geom_treemap() +
  geom_treemap_subgroup_border(color = "white", size = 3) +
  geom_treemap_subgroup_text(
    place = "center",
    grow = TRUE,
    alpha = 0.5,
    color = "black",
    fontface = "bold",
    size = 18
  ) +
  geom_treemap_text(
    color = "white",
    place = "center",
    size = 10,
    grow = FALSE
  ) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Patient Distribution by Region and Treatment",
       subtitle = "Rectangle size proportional to patient count",
       fill = "Region") +
  theme_custom() +
  theme(legend.position = "right")
```

## Interpreting Treemaps

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**Key Features:**

- **Size = Quantity** - Larger rectangles = more patients
- **Color = Category** - Each region has its own color
- **Hierarchy** - Region (outer) → Treatment (inner)
- **Space-efficient** - Uses all available area

**Best Practices:**

- Limit to 2-3 hierarchy levels
- Use clear labels
- Consistent color scheme
- Consider interactivity for complex trees
:::

::: {.column width="50%"}
**Our Treemap Shows:**

1. **East region largest** overall
   - Most patients

2. **Within each region** treatment distribution visible
   - Relative sizes easy to compare

3. **Quick identification** of smallest/largest segments

4. **Hierarchical structure** immediately clear
   - Region → Treatment breakdown

**Advantage:** Efficient use of space compared to nested pie charts
:::
:::

## Sunburst Chart - Code

::: {style="font-size: 0.75em;"}
**Purpose:** Radial hierarchical visualization

**Goal:** 🔴 Part of Whole (circular)

```{r}
#| echo: true
#| eval: false

# Prepare data in required format for sunburst
# Format: path-count pairs
sunburst_data <- health_data %>%
  filter(!is.na(Region), !is.na(Gender), !is.na(SmokingStatus)) %>%
  count(Region, Gender, SmokingStatus) %>%
  mutate(path = paste(Region, Gender, SmokingStatus, sep = "-")) %>%
  select(path, n)

# Create sunburst chart
sunburst(
  data = sunburst_data,
  count = TRUE,
  legend = list(w = 150, h = 25, s = 5, t = 25),
  breadcrumb = list(w = 0, h = 25, s = 5, t = 10),
  colors = list(
    range = c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
             "#FF7F00", "#FFFF33", "#A65628", "#F781BF")
  ),
  withD3 = TRUE,
  width = "100%",
  height = 500
)
```

**Note:** Sunburst creates interactive HTML widget. For static, use `ggplot2` alternative.

**Static Alternative (next slide):**
:::

## Sunburst Alternative: Radial Bar - Code

::: {style="font-size: 0.8em;"}
```{r}
#| echo: true
#| eval: false

# Create radial bar plot (sunburst alternative)
radial_data <- health_data %>%
  filter(!is.na(Region), !is.na(SmokingStatus)) %>%
  count(Region, SmokingStatus) %>%
  group_by(Region) %>%
  mutate(
    percentage = n / sum(n) * 100,
    id = paste(Region, SmokingStatus, sep = "-")
  )

ggplot(radial_data, aes(x = Region, y = n, fill = SmokingStatus)) +
  geom_col(width = 1, color = "white", size = 0.5) +
  coord_polar() +
  scale_fill_manual(values = c("Never" = "#00A087", 
                              "Former" = "#3C5488",
                              "Current" = "#DC0000")) +
  labs(title = "Smoking Status Distribution by Region (Radial)",
       subtitle = "Each segment shows proportion within region",
       fill = "Smoking Status") +
  theme_custom() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```
:::

## Radial Bar - Output

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

radial_data <- health_data %>%
  filter(!is.na(Region), !is.na(SmokingStatus)) %>%
  count(Region, SmokingStatus) %>%
  group_by(Region) %>%
  mutate(
    percentage = n / sum(n) * 100,
    id = paste(Region, SmokingStatus, sep = "-")
  )

ggplot(radial_data, aes(x = Region, y = n, fill = SmokingStatus)) +
  geom_col(width = 1, color = "white", size = 0.5) +
  coord_polar() +
  scale_fill_manual(values = c("Never" = "#00A087", 
                              "Former" = "#3C5488",
                              "Current" = "#DC0000")) +
  labs(title = "Smoking Status Distribution by Region (Radial)",
       subtitle = "Each segment shows proportion within region",
       fill = "Smoking Status") +
  theme_custom() +
  theme(
    axis.text.y = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )
```

## Marimekko Chart - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Two-way composition (varying column widths)

**Goal:** 🔴 Part of Whole (two dimensions)

```{r}
#| echo: true
#| eval: false

# Prepare data for Marimekko
marimekko_data <- health_data %>%
  filter(!is.na(Region), !is.na(TreatmentGroup)) %>%
  count(Region, TreatmentGroup) %>%
  group_by(Region) %>%
  mutate(
    region_total = sum(n),
    pct_in_region = n / region_total * 100
  ) %>%
  ungroup() %>%
  mutate(
    region_pct = region_total / sum(region_total) * 100,
    cumsum_region = cumsum(region_pct) - region_pct / 2
  )

# Create Marimekko chart
ggplot(marimekko_data, aes(x = Region, y = pct_in_region, 
                           fill = TreatmentGroup, width = region_pct)) +
  geom_col(position = "stack", color = "white", size = 0.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Treatment Distribution by Region (Marimekko)",
       subtitle = "Column width = regional patient share | Height = treatment proportion",
       x = "Region",
       y = "Percentage within Region",
       fill = "Treatment") +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 0))
```
:::

## Marimekko Chart - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false

marimekko_data <- health_data %>%
  filter(!is.na(Region), !is.na(TreatmentGroup)) %>%
  count(Region, TreatmentGroup) %>%
  group_by(Region) %>%
  mutate(
    region_total = sum(n),
    pct_in_region = n / region_total * 100
  ) %>%
  ungroup() %>%
  mutate(
    region_pct = region_total / sum(region_total) * 100
  )

ggplot(marimekko_data, aes(x = Region, y = pct_in_region, 
                           fill = TreatmentGroup)) +
  geom_col(position = "stack", color = "white", size = 0.5) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Treatment Distribution by Region (Marimekko)",
       subtitle = "Height shows treatment proportion within each region",
       x = "Region",
       y = "Percentage within Region",
       fill = "Treatment") +
  theme_custom()
```

## Interpreting Marimekko Charts

::: {style="font-size: 0.9em;"}
**Unique Features:**

::: {.incremental}
1. **Two dimensions of composition**
   - Width would represent one variable's total
   - Height shows breakdown within that category

2. **Space = Area**
   - Total area of each segment = actual count contribution

3. **Compare both dimensions simultaneously**
   - More information dense than standard stacked bars

**Our Chart:**

- **Treatment distribution** fairly consistent across regions
- **All regions** show similar proportions of each treatment
- **Visual balance** suggests good randomization

**When to use:** Two categorical variables where both totals and proportions matter
:::
:::

## Waffle Chart - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Unit-based visualization (each square = fixed count)

**Goal:** 🔴 Part of Whole (intuitive units)

```{r}
#| echo: true
#| eval: false


# Prepare data (each unit = 10 patients)
waffle_data <- health_data %>%
  filter(!is.na(TreatmentGroup)) %>%
  count(TreatmentGroup) %>%
  mutate(squares = round(n / 10))  # 1 square = 10 patients

# Create waffle chart
waffle_expanded <- waffle_data %>%
  rowwise() %>%
  mutate(id = list(1:squares)) %>%
  unnest(id) %>%
  mutate(square_id = row_number(),
         x = (square_id - 1) %% 10,   # 10 rows
         y = (square_id - 1) %/% 10)

# Plot with ggplot2
ggplot(waffle_expanded, aes(x = x, y = y, fill = TreatmentGroup)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_manual(values = c("Treatment A" = "#E41A1C",
                               "Treatment B" = "#377EB8",
                               "Control"     = "#4DAF4A")) +
  coord_equal() +
  labs(title = "Patient Distribution by Treatment Group",
       x = "1 square = 10 patients",
       y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

```
:::

## Waffle Chart - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
#| message: false

waffle_data <- health_data %>%
  filter(!is.na(TreatmentGroup)) %>%
  count(TreatmentGroup) %>%
  mutate(squares = round(n / 10))

# Manual waffle using geom_tile
waffle_expanded <- waffle_data %>%
  rowwise() %>%
  mutate(id = list(1:squares)) %>%
  unnest(id) %>%
  mutate(square_id = row_number(),
         x = (square_id - 1) %% 10,   # 10 rows
         y = (square_id - 1) %/% 10)

# Plot with ggplot2
ggplot(waffle_expanded, aes(x = x, y = y, fill = TreatmentGroup)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_manual(values = c("Treatment A" = "#E41A1C",
                               "Treatment B" = "#377EB8",
                               "Control"     = "#4DAF4A")) +
  coord_equal() +
  labs(title = "Patient Distribution by Treatment Group",
       x = "1 square = 10 patients",
       y = NULL) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
  )

```

## Interpreting Waffle Charts

::: {style="font-size: 0.9em;"}
**Why Waffle Charts Work:**

::: {.incremental}
1. **Intuitive unit representation**
   - Each square = fixed count (e.g., 10 patients)
   - Easy to count and compare

2. **Avoid pie chart problems**
   - No angles to judge
   - Better for exact comparisons

3. **Visual impact**
   - Engaging and memorable
   - Good for presentations and infographics

**Our Chart:**

- **Treatment groups balanced**
- **Control group** has similar size
- **Each square represents 10 patients**
- **Visual counting** easier than interpreting angles

**When to use:** Small to medium datasets, need visual impact
:::
:::

## Dendrogram - Code

::: {style="font-size: 0.75em;"}
**Purpose:** Show hierarchical clustering of categories

**Goal:** ⚫ Correlation + hierarchical structure

```{r}
#| echo: true
#| eval: false

# Calculate dissimilarity matrix for regions based on health metrics
region_profiles <- health_data %>%
  filter(!is.na(Region)) %>%
  group_by(Region) %>%
  summarise(
    mean_bp = mean(SysBP_0m, na.rm = TRUE),
    mean_chol = mean(Cholesterol_0m, na.rm = TRUE),
    mean_bmi = mean(BMI_0m, na.rm = TRUE),
    pct_smokers = mean(SmokingStatus == "Current", na.rm = TRUE) * 100
  ) %>%
  column_to_rownames("Region")

# Scale variables
region_scaled <- scale(region_profiles)

# Hierarchical clustering
hc <- hclust(dist(region_scaled), method = "ward.D2")

# Plot dendrogram
library(ggdendro)
ggdendrogram(hc, rotate = FALSE, theme_dendro = FALSE) +
  labs(title = "Hierarchical Clustering of Regions by Health Profiles",
       subtitle = "Based on BP, cholesterol, BMI, and smoking rates",
       x = "Region",
       y = "Distance (Ward's method)") +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 0))
```
:::

## Dendrogram - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
#| message: false

library(ggdendro)
library(tibble)

region_profiles <- health_data %>%
  filter(!is.na(Region)) %>%
  group_by(Region) %>%
  summarise(
    mean_bp = mean(SysBP_0m, na.rm = TRUE),
    mean_chol = mean(Cholesterol, na.rm = TRUE),
    mean_bmi = mean(BMI, na.rm = TRUE),
    pct_smokers = mean(SmokingStatus == "Current", na.rm = TRUE) * 100
  ) %>%
  column_to_rownames("Region")

region_scaled <- scale(region_profiles)
hc <- hclust(dist(region_scaled), method = "ward.D2")

ggdendrogram(hc, rotate = FALSE, theme_dendro = FALSE) +
  labs(title = "Hierarchical Clustering of Regions by Health Profiles",
       subtitle = "Based on BP, cholesterol, BMI, and smoking rates",
       x = "Region",
       y = "Distance (Ward's method)") +
  theme_custom() +
  theme(axis.text.x = element_text(angle = 0))
```

## Interpreting Dendrograms

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**Reading Dendrograms:**

- **Height** = Dissimilarity
- **Branches** joining lower = more similar
- **Clusters** formed by cutting at height
- **Leaf** = Individual category

**Method Matters:**

- **Ward's** - Minimizes within-cluster variance
- **Complete** - Maximum distance
- **Average** - Mean distance between clusters
:::

::: {.column width="50%"}
**Our Dendrogram:**

1. **East and North** most similar
   - Join at lowest height
   - Similar health profiles

2. **South different** from North/East
   - Joins at higher level

3. **West most distinct**
   - Joins last, highest dissimilarity

**Use Case:** Identifying similar market segments, patient groups
:::
:::

## Chart Selection Guide

::: {style="font-size: 0.85em;"}
| Chart Type | Goal | Best For | Complexity |
|------------|------|----------|------------|
| **Treemap** | 🔴 Part of Whole | Hierarchical proportions, space-efficient | Medium |
| **Sunburst** | 🔴 Part of Whole | Multi-level hierarchies, interactive | High |
| **Marimekko** | 🔴 Part of Whole | Two-way composition | Medium-High |
| **Waffle** | 🔴 Part of Whole | Small datasets, visual impact | Easy-Medium |
| **Dendrogram** | ⚫ Correlation | Category clustering, similarity | Medium-High |

**Pro Tip:** Advanced charts require more explanation - include clear legends and labels
:::

## Common Mistakes

::: {.incremental style="font-size: 0.9em;"}
1. **Too many categories** in treemap/sunburst
   - Solution: Filter to top N categories or aggregate small ones

2. **Poor color choices** making comparisons hard
   - Solution: Use consistent, meaningful color schemes

3. **Missing context** in dendrograms
   - Solution: Explain clustering method and distance metric

4. **Waffle charts** with inappropriate units
   - Solution: Choose unit size that creates 50-200 squares

5. **Marimekko** without clear labels
   - Solution: Add annotations explaining width/height meanings

6. **Interactive charts** not working in static reports
   - Solution: Provide static alternatives or export images
:::

## Practice Exercise

**Your Turn:**

1. Create a treemap showing patient distribution by Region → Gender → SmokingStatus
2. Build a waffle chart for ExerciseFreq (1 square = 5 patients)
3. Create a dendrogram clustering treatment groups based on outcome metrics
4. Make a Marimekko chart for Gender × TreatmentGroup

**Bonus Challenge:**

Create an interactive sunburst chart with 3 levels: Region → TreatmentGroup → Adherence_Status

## Summary

::: {style="font-size: 0.9em;"}
**What You've Learned:**

::: {.incremental}
1. **Treemaps** efficiently show hierarchical proportions
2. **Sunburst charts** display circular hierarchies
3. **Marimekko charts** reveal two-way compositions
4. **Waffle charts** provide intuitive unit-based views
5. **Dendrograms** identify category similarities

**Key Principle:**

> Advanced categorical charts reveal patterns in complex, multi-level categorical data that simpler charts cannot show
:::
:::

## Next Steps {.center}

**You've mastered Categorical - Advanced!**

**Next:** Continuous - Advanced

Learn advanced continuous visualizations:

- Contour plots
- 3D surface plots
- Parallel coordinates
- Advanced density techniques
