---
title: "Continuous - ADVANCED"
subtitle: "Multivariate & 3D Techniques"
format:
  revealjs:
    theme: serif
    slide-number: true
    code-fold: true
---

# Continuous Advanced {.center}

## Topics Covered

- Contour plots (2D density)
- 3D surface plots
- Parallel coordinates
- Correlation ellipses
- Hexbin plots (large data)

Goal: Handle complex continuous relationships

## Setup

```{r}
library(readxl); library(dplyr); library(ggplot2)
library(plotly); library(GGally)
health_data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")
```

## Contour Plot - Code

```{r eval=FALSE, echo=TRUE}
ggplot(health_data, aes(x = BMI_0m, y = SysBP_0m)) +
  geom_density_2d_filled(alpha = 0.7) +
  geom_point(size = 0.5, alpha = 0.3) +
  scale_fill_viridis_d(option = "plasma") +
  labs(title = "2D Density: BMI vs Blood Pressure",
       x = "BMI", y = "Systolic BP") +
  theme_minimal()
```

Shows concentration areas - darker = more patients

## Parallel Coordinates - Code

```{r eval=FALSE, echo=TRUE}
library(GGally)
health_subset <- health_data %>%
  select(Age, BMI_0m, SysBP_0m, Cholesterol_0m, BloodSugar_0m) %>%
  na.omit() %>% sample_n(200)

ggparcoord(health_subset, columns = 1:5,
           scale = "uniminmax",
           alphaLines = 0.3) +
  labs(title = "Parallel Coordinates: Health Metrics") +
  theme_minimal()
```

Each line = one patient across multiple variables

## 3D Surface (Conceptual)

```{r eval=FALSE, echo=TRUE}
library(plotly)
# Create grid
x <- seq(20, 40, 0.5)
y <- seq(100, 180, 2)
# Simulate density surface
z <- outer(x, y, function(a,b) 
  dnorm(a, 28, 5) * dnorm(b, 140, 20))

plot_ly(x = ~x, y = ~y, z = ~z) %>%
  add_surface() %>%
  layout(title = "3D Density Surface: BMI × BP")
```

Interactive 3D exploration of relationships

## Hexbin Plot - Code

```{r eval=FALSE, echo=TRUE}
ggplot(health_data, aes(x = BMI_0m, y = SysBP_0m)) +
  geom_hex(bins = 30) +
  scale_fill_viridis_c(option = "inferno", trans = "log10") +
  labs(title = "Hexbin Plot: Large Dataset Visualization",
       fill = "Count (log10)") +
  theme_minimal()
```

Better than scatter for 1000+ points

## Correlation Matrix with Ellipses

```{r eval=FALSE, echo=TRUE}
library(ggcorrplot)
cor_matrix <- health_data %>%
  select(Age, BMI_0m, SysBP_0m, DiaBP_0m, 
         Cholesterol_0m, BloodSugar_0m) %>%
  cor(use = "complete.obs")

ggcorrplot(cor_matrix, type = "lower",
           lab = TRUE, lab_size = 3,
           colors = c("#DC0000", "white", "#00A087"))
```

Quick overview of all correlations

## Summary

**Advanced continuous techniques:**

1. Contour plots - 2D density
2. Parallel coordinates - multivariate patterns  
3. 3D surfaces - interactive exploration
4. Hexbin - big data visualization
5. Correlation ellipses - relationship matrices

**Next:** Category vs Continuous Advanced
