---
title: "Time Series - ADVANCED"
subtitle: "Forecasting & Decomposition"
format:
  revealjs:
    theme: serif
    slide-number: true
    code-fold: true
---

# Time Series Advanced {.center}

## Topics Covered

- STL decomposition (trend/seasonal)
- Forecasting with ARIMA
- Prophet models
- Change point detection
- Spectral analysis (FFT)

Goal: Advanced temporal patterns

## Setup

```{r}
library(readxl); library(tidyverse); library(ggplot2)
library(forecast); library(prophet)
health_data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Create time series
ts_data <- health_data %>%
  pivot_longer(c(SysBP_0m, SysBP_3m, SysBP_6m),
              names_to = "Time", values_to = "BP") %>%
  group_by(Time, TreatmentGroup) %>%
  summarise(mean_bp = mean(BP, na.rm = TRUE))
```

## STL Decomposition - Code

```{r eval=FALSE, echo=TRUE}
# Create proper time series (conceptual with more timepoints)
bp_ts <- ts(monthly_bp_data$mean_bp, frequency = 12)

# STL decomposition
stl_result <- stl(bp_ts, s.window = "periodic")

# Plot components
autoplot(stl_result) +
  labs(title = "STL Decomposition of Blood Pressure",
       subtitle = "Trend + Seasonal + Remainder") +
  theme_minimal()
```

Separates trend, seasonality, and noise

## Forecasting - Code

```{r eval=FALSE, echo=TRUE}
# Fit ARIMA model
arima_model <- auto.arima(bp_ts)

# Forecast 12 months ahead
forecast_result <- forecast(arima_model, h = 12)

# Plot
autoplot(forecast_result) +
  labs(title = "12-Month Blood Pressure Forecast",
       y = "Mean Systolic BP") +
  theme_minimal()
```

Blue band = 80% CI, gray = 95% CI

## Prophet Forecasting

```{r eval=FALSE, echo=TRUE}
library(prophet)
# Prepare data for Prophet
prophet_data <- data.frame(
  ds = as.Date(monthly_bp_data$date),
  y = monthly_bp_data$mean_bp
)

# Fit and predict
m <- prophet(prophet_data)
future <- make_future_dataframe(m, periods = 12, freq = "month")
forecast <- predict(m, future)

# Plot
plot(m, forecast) +
  labs(title = "Prophet Forecast with Trend")
```

Handles holidays, changepoints automatically

## Change Point Detection

```{r eval=FALSE, echo=TRUE}
library(changepoint)
# Detect change points in mean
cpt_result <- cpt.mean(bp_ts, method = "PELT")

# Plot
plot(cpt_result, cpt.col = "red", cpt.width = 3,
    main = "Change Point Detection in BP Time Series")
```

Identifies when trend shifts significantly

## Rolling Statistics Plot

```{r eval=FALSE, echo=TRUE}
library(zoo)
# Calculate rolling mean and SD
rolling_data <- monthly_bp_data %>%
  mutate(
    roll_mean = rollmean(mean_bp, k = 3, fill = NA),
    roll_sd = rollapply(mean_bp, width = 3, FUN = sd,
                       fill = NA)
  )

ggplot(rolling_data, aes(x = date)) +
  geom_line(aes(y = mean_bp), color = "gray70") +
  geom_line(aes(y = roll_mean), color = "blue", size = 1) +
  geom_ribbon(aes(ymin = roll_mean - roll_sd,
                 ymax = roll_mean + roll_sd),
             alpha = 0.2, fill = "blue") +
  labs(title = "3-Month Rolling Mean ± SD") +
  theme_minimal()
```

## Summary

**Advanced time series:**

1. STL decomposition - separate components
2. ARIMA forecasting - statistical predictions
3. Prophet - flexible, interpretable forecasts
4. Change points - detect regime shifts
5. Rolling statistics - smoothed trends

**Next:** Network Advanced
