---
title: "Practical Application"
subtitle: "Real-World Case Studies"
format:
  revealjs:
    theme: serif
    css: ../assets/slide_deck.css
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# From Analysis to Action {.center}

## Case Study 1: Publication Figure

**Scenario:** Need publication-ready figure for medical journal

**Requirements:**
- High resolution (300 DPI)
- Grayscale-friendly
- Clear statistical annotations
- Professional formatting

## Publication Figure - Code

```{r eval=FALSE, echo=TRUE}
# Treatment effect with confidence intervals
model <- lm(SysBP_12m ~ TreatmentGroup + Age + BMI_0m + SysBP_0m,
           data = health_data)

library(ggeffects)
effects <- ggpredict(model, terms = "TreatmentGroup")

p <- plot(effects) +
  labs(title = "Treatment Effect on Systolic Blood Pressure",
       subtitle = "Adjusted for age, BMI, and baseline BP",
       x = "Treatment Group",
       y = "Predicted Systolic BP (mmHg)",
       caption = "Error bars: 95% confidence intervals. n=500 per group.") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0),
    panel.grid.minor = element_blank(),
    legend.position = "none"
  )

# Save high-res
ggsave("figure1.tiff", p, width = 7, height = 5, dpi = 300)
ggsave("figure1.eps", p, width = 7, height = 5, device = cairo_ps)
```

## Case Study 2: Executive Dashboard

**Scenario:** Monthly report for C-suite

**Requirements:**
- Key metrics at a glance
- Trend over time
- Regional breakdown
- Action items highlighted

## Executive Dashboard Structure

```{r eval=FALSE, echo=TRUE}
---
title: "Q4 2024 Health Outcomes"
output:
  flexdashboard::flex_dashboard:
    theme: cosmo
---

# Executive Summary {data-icon="fa-dashboard"}

## Row {data-height=150}

### Total Patients
valueBox(2500, icon = "fa-users", color = "primary")

### Avg BP Reduction  
valueBox(-15.3, caption = "mmHg", icon = "fa-arrow-down",
        color = "success")

### Target Achievement
gauge(87, min = 0, max = 100, symbol = "%",
     gaugeSectors(success = c(80, 100), warning = c(60, 80),
                 danger = c(0, 60)))

## Row

### Trend Over Time
(plot showing quarterly BP trends)

### Regional Heatmap
(choropleth of outcomes by region)
```

## Case Study 3: Exploratory Analysis Report

**Scenario:** Initial data exploration for research team

**Requirements:**
- Comprehensive EDA
- Identify patterns
- Flag data quality issues
- Generate hypotheses

## EDA Report Template

```{r eval=FALSE, echo=TRUE}
---
title: "Exploratory Data Analysis: BP Study"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# Data Overview
- Summary statistics
- Missing data patterns
- Outlier detection

# Univariate Analysis
- Distribution of each variable
- Histograms, density plots

# Bivariate Analysis
- Correlation matrix
- Key relationships

# Multivariate Patterns
- PCA or t-SNE
- Clustering analysis

# Data Quality Issues
- Missing data by group
- Potential errors
- Recommendations
```

## Case Study 4: Interactive Web App

**Scenario:** Tool for clinicians to explore patient data

**Requirements:**
- Filter by patient characteristics
- Show similar patients
- Display trends
- Export reports

## Clinician Tool - Shiny

```{r eval=FALSE, echo=TRUE}
ui <- fluidPage(
  titlePanel("Patient Explorer"),
  sidebarLayout(
    sidebarPanel(
      textInput("patientID", "Enter Patient ID:"),
      selectInput("metric", "Compare:",
                 choices = c("BP", "Cholesterol", "BMI")),
      actionButton("find", "Find Similar Patients")
    ),
    mainPanel(
      tabsetPanel(
        tabPanel("Patient Profile",
                valueBoxOutput("currentBP"),
                plotOutput("patientTrend")),
        tabPanel("Similar Patients",
                plotOutput("similarScatter"),
                DTOutput("similarTable")),
        tabPanel("Predictions",
                plotOutput("forecastPlot"))
      )
    )
  )
)

server <- function(input, output) {
  # Reactive logic for patient comparison
  # Machine learning for similarity
  # Visualization of results
}
```

## Case Study 5: Automated Reporting

**Scenario:** Weekly reports for 50+ clinic sites

**Requirements:**
- Automated generation
- Site-specific content
- Consistent formatting
- Email distribution

## Parameterized Reports

```{r eval=FALSE, echo=TRUE}
---
title: "Weekly Report: `r params$site_name`"
output: html_document
params:
  site_name: "North Clinic"
  week_start: "2024-01-01"
  week_end: "2024-01-07"
---


# Filter data for this site
site_data <- health_data %>%
  filter(Site == params$site_name,
        Date >= params$week_start,
        Date <= params$week_end)

# Generate site-specific visualizations
```

# Generate for all sites
```{r eval=FALSE}
sites <- unique(health_data$Site)
for (site in sites) {
  rmarkdown::render("template.Rmd",
                   params = list(site_name = site),
                   output_file = paste0(site, "_report.html"))
}
```

## Workflow Best Practices

**Project Organization:**

```
project/
├── data/
│   ├── raw/
│   └── processed/
├── R/
│   ├── 01_cleaning.R
│   ├── 02_analysis.R
│   └── functions.R
├── output/
│   ├── figures/
│   └── tables/
├── reports/
│   └── analysis.Rmd
└── README.md
```

**Version Control:**
- Use Git for code
- Document changes
- Reproducible workflows

## Code Quality Checklist

**Before Sharing:**

- [ ] Remove hardcoded paths
- [ ] Add comments to complex code
- [ ] Check for errors/warnings
- [ ] Test with subset of data
- [ ] Verify output matches expectations
- [ ] Include session info (`sessionInfo()`)
- [ ] Document package versions (`renv`)

## Summary

**Real-world skills:**

1. Publication figures - high-quality, standards-compliant
2. Executive dashboards - key metrics, trends
3. EDA reports - comprehensive exploration
4. Interactive tools - Shiny apps for users
5. Automated reporting - scale insights

**Remember:**
- Know your audience
- Tell a clear story
- Make it reproducible
- Test thoroughly

## Congratulations! 🎉

**You've completed all levels:**

- ✓ Foundation
- ✓ Basic (4 data types)
- ✓ Intermediate (6 techniques)
- ✓ Advanced (7 specialized topics)
- ✓ Practical application

**You can now:**
- Choose appropriate visualizations
- Create publication-quality figures
- Build interactive dashboards
- Apply advanced techniques
- Handle real-world scenarios

**Keep practicing and exploring!**
