---
title: "Categorical Plots - INTERMEDIATE"
subtitle: "Enhanced Categorical Visualizations"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Categorical Intermediate {.center}

## Enhanced Categorical Visualizations

::: {style="font-size: 0.9em;"}
**Building on Basic Skills:**

Moving beyond simple bar and pie charts to professional-grade categorical visualizations.

**New Techniques:**
- Lollipop charts (cleaner than bars)
- Cleveland dot plots (compare groups)
- Mosaic plots (show associations)
- Radial/circular charts (visual appeal)

**New Packages:**
- `forcats` for factor manipulation
- `ggmosaic` for mosaic plots
- Enhanced `ggplot2` techniques
:::

## Setup

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-5|7-8|10-12"

# Load required libraries
library(ggplot2)
library(dplyr)
library(forcats)     # For factor reordering
library(ggmosaic)    # For mosaic plots

# Load dataset
data <- readxl::read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Clean data: remove missing values in key variables
data_clean <- data %>%
  filter(!is.na(ExerciseFreq), !is.na(ChronicDisease))
```

## Data Preview {.smaller}

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(forcats)
library(ggmosaic)
library(readxl)

data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")
data_clean <- data %>% filter(!is.na(ExerciseFreq), !is.na(ChronicDisease))

data_clean %>% 
  count(Region, SmokingStatus) %>%
  head(10)
```

---

# Lollipop Chart

## Lollipop Chart - Code

🟢 **Goal:** Ranking (cleaner alternative to bars)

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-4|6-9|10-11|12-14"

# Calculate counts and prepare for ordered lollipop
region_counts <- data_clean %>%
  count(Region) %>%
  arrange(n)  # Sort for proper ordering

# Create lollipop chart
ggplot(region_counts, aes(x = n, y = fct_reorder(Region, n))) +
  geom_segment(aes(x = 0, xend = n, y = Region, yend = Region),
               color = "gray70", size = 1.5) +  # Draw lollipop sticks
  geom_point(size = 5, color = "#2C5F8D") +  # Draw lollipop heads
  geom_text(aes(label = n), hjust = -0.5, family = "serif", size = 4) +  # Add count labels
  labs(title = "Patient Count by Region (Lollipop Chart)",
       x = "Number of Patients", y = "Region") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        panel.grid.major.y = element_blank())  # Remove horizontal grid
```

## Lollipop Chart - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

region_counts <- data_clean %>%
  count(Region) %>%
  arrange(n)

ggplot(region_counts, aes(x = n, y = fct_reorder(Region, n))) +
  geom_segment(aes(x = 0, xend = n, y = Region, yend = Region),
               color = "gray70", size = 1.5) +
  geom_point(size = 5, color = "#2C5F8D") +
  geom_text(aes(label = n), hjust = -0.5, family = "serif", size = 4) +
  labs(title = "Patient Count by Region (Lollipop Chart)",
       x = "Number of Patients", y = "Region") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        panel.grid.major.y = element_blank())
```

---

# Interpreting Lollipop Charts

## Reading Lollipop and Dot Plots

**Using the previous plot:**

::: {.incremental style="font-size: 0.85em;"}
**OBSERVATION:**
- East region: ~40 patients (lowest)
- West and South: ~45 patients each (middle tier)
- North region: ~65 patients (highest, ~60% more than East)
- Clear ranking from bottom to top
- Relatively balanced except North dominance

**INTERPRETATION:**
- North region significantly overrepresented
- East region undersampled  
- West and South balanced
- May reflect population distribution or recruitment success
- Geographic imbalance could affect generalizability

**IMPLICATIONS:**
- Regional stratification needed in analysis
- Results may skew toward North region characteristics
- Consider weighting by region for national estimates
- Recruitment challenges in East region

**ACTIONS:**
- Include region as covariate in models
- Report results stratified by region
- Consider regional subgroup analyses
- Investigate reasons for North oversampling
:::
:::

---

# Cleveland Dot Plot

## Cleveland Dot Plot - Code

🟢 **Goal:** Compare values across two groups

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-5|7-11|12-14|15-17"

# Calculate mean BMI by smoking status and gender
bmi_summary <- data_clean %>%
  filter(Gender %in% c("Male", "Female")) %>%  # Focus on main genders
  group_by(SmokingStatus, Gender) %>%
  summarise(Mean_BMI = mean(BMI, na.rm = TRUE), .groups = "drop")

# Create Cleveland dot plot
ggplot(bmi_summary, aes(x = Mean_BMI, y = fct_reorder(SmokingStatus, Mean_BMI),
                        color = Gender, group = Gender)) +
  geom_line(aes(group = SmokingStatus), color = "gray80", size = 1.5) +  # Connect dots
  geom_point(size = 5) +  # Add colored dots
  scale_color_manual(values = c("Female" = "#E91E63", "Male" = "#2196F3"),
                     name = "Gender") +
  labs(title = "Mean BMI by Smoking Status and Gender",
       x = "Mean BMI", y = "Smoking Status") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12))
```

## Cleveland Dot Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false

bmi_summary <- data_clean %>%
  filter(Gender %in% c("Male", "Female")) %>%
  group_by(SmokingStatus, Gender) %>%
  summarise(Mean_BMI = mean(BMI, na.rm = TRUE), .groups = "drop")

ggplot(bmi_summary, aes(x = Mean_BMI, y = fct_reorder(SmokingStatus, Mean_BMI),
                        color = Gender, group = Gender)) +
  geom_line(aes(group = SmokingStatus), color = "gray80", size = 1.5) +
  geom_point(size = 5) +
  scale_color_manual(values = c("Female" = "#E91E63", "Male" = "#2196F3"),
                     name = "Gender") +
  labs(title = "Mean BMI by Smoking Status and Gender",
       x = "Mean BMI", y = "Smoking Status") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Mosaic Plot

## Mosaic Plot - Code

🔴 **Goal:** Part of Whole + Show categorical associations

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3-5|6-7|8-10"

# Create mosaic plot showing relationship between Region and Treatment Group
ggplot(data = data_clean) +
  geom_mosaic(aes(x = product(TreatmentGroup, Region), 
                  fill = TreatmentGroup)) +  # Create mosaic tiles
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB"),
                    name = "Treatment Group") +
  labs(title = "Treatment Group Distribution by Region (Mosaic Plot)",
       x = "Region", y = "Treatment Group Proportion") +
  theme_mosaic(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text.x = element_text(angle = 0))
```

::: {.notes}
In mosaic plots, area = proportion. Width = Region proportion, Height = Treatment proportion within region
:::

## Mosaic Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data = data_clean) +
  geom_mosaic(aes(x = product(TreatmentGroup, Region), 
                  fill = TreatmentGroup)) +
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB"),
                    name = "Treatment Group") +
  labs(title = "Treatment Group Distribution by Region (Mosaic Plot)",
       x = "Region", y = "Treatment Group Proportion") +
  theme_mosaic(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Interpreting Mosaic Plots

## How to Read Mosaic Plots

**Using the previous plot:**

::: {.incremental style="font-size: 0.8em;"}
**Key Principle:** Area = Proportion (not just height or width!)

**OBSERVATION:**
- North region: Widest column (~40% of total width) = most patients
- All regions: Treatment groups roughly 50-50 split (equal heights within columns)
- Tiles aligned horizontally = no strong association
- East region: Narrowest column (~20% width) = fewest patients

**INTERPRETATION:**
- **Width of columns:** Regional sample sizes (North > South ≈ West > East)
- **Height within columns:** Treatment allocation within each region
- **Alignment:** Horizontal edges align = random assignment successful
- **No association:** Treatment group independent of region (good randomization!)

**IMPLICATIONS:**
- Randomization worked well across all regions
- No geographic bias in treatment allocation
- Can analyze regions separately without treatment imbalance concern
- Balanced design supports causal inference

**Reading Mosaic Tips:**
- Misaligned tiles = association exists
- Aligned tiles = independence (no association)
- Area, not just height, represents proportion
- Compare conditional distributions (within columns)
:::
:::

---

# Mosaic Plot - Three Variables

## Complex Mosaic - Code

🔴 **Goal:** Multi-way categorical analysis

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-3|4-7|8-10"

# Three-way mosaic: Gender × Smoking × Treatment
ggplot(data = data_clean %>% filter(Gender %in% c("Male", "Female"))) +
  geom_mosaic(aes(x = product(SmokingStatus, Gender, TreatmentGroup), 
                  fill = SmokingStatus),
              divider = ddecker()) +  # Use double-decker layout
  scale_fill_manual(values = c("Current smoker" = "#E74C3C",
                                "Former smoker" = "#F39C12",
                                "Non-smoker" = "#27AE60"),
                    name = "Smoking Status") +
  labs(title = "Smoking Status by Gender and Treatment Group",
       x = "Gender : Treatment Group") +
  theme_mosaic(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9))
```

## Complex Mosaic - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data = data_clean %>% filter(Gender %in% c("Male", "Female"))) +
  geom_mosaic(aes(x = product(SmokingStatus, Gender, TreatmentGroup), 
                  fill = SmokingStatus),
              divider = ddecker()) +
  scale_fill_manual(values = c("Current smoker" = "#E74C3C",
                                "Former smoker" = "#F39C12",
                                "Non-smoker" = "#27AE60"),
                    name = "Smoking Status") +
  labs(title = "Smoking Status by Gender and Treatment Group",
       x = "Gender : Treatment Group") +
  theme_mosaic(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Radial Bar Chart

## Radial Bar Chart - Code

🟢 **Goal:** Ranking with visual appeal (use sparingly!)

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-4|6-8|9-11|12-15"

# Prepare data for radial plot
exercise_counts <- data_clean %>%
  count(ExerciseFreq) %>%
  mutate(ExerciseFreq = fct_reorder(ExerciseFreq, n))

# Create radial bar chart
ggplot(exercise_counts, aes(x = ExerciseFreq, y = n, fill = ExerciseFreq)) +
  geom_col(alpha = 0.8, color = "white", size = 1) +  # Create bars
  coord_polar(theta = "y") +  # Transform to circular coordinates
  scale_fill_brewer(palette = "Set2", name = "Exercise Frequency") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            family = "serif", size = 5, fontface = "bold") +  # Add labels
  labs(title = "Exercise Frequency Distribution (Radial)",
       x = NULL, y = NULL) +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank())
```

## Radial Bar Chart - Plot

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 7

exercise_counts <- data_clean %>%
  count(ExerciseFreq) %>%
  mutate(ExerciseFreq = fct_reorder(ExerciseFreq, n))

ggplot(exercise_counts, aes(x = ExerciseFreq, y = n, fill = ExerciseFreq)) +
  geom_col(alpha = 0.8, color = "white", size = 1) +
  coord_polar(theta = "y") +
  scale_fill_brewer(palette = "Set2", name = "Exercise Frequency") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            family = "serif", size = 5, fontface = "bold") +
  labs(title = "Exercise Frequency Distribution (Radial)",
       x = NULL, y = NULL) +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid = element_blank(),
        legend.position = "bottom",
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Quick Reference

## Categorical Intermediate - Summary

| Chart Type | Goal | When to Use | Code Lines |
|------------|------|-------------|-----------|
| **Lollipop** | 🟢 Ranking | Cleaner than bars, 5-20 categories | 10-15 |
| **Cleveland Dot** | 🟢 Compare Groups | Show differences between 2 groups | 10-15 |
| **Mosaic (2-way)** | 🔴 Part + Association | Test independence of 2 categorical variables | 10-15 |
| **Mosaic (3-way)** | 🔴 Multi-way | Complex categorical relationships | 15-20 |
| **Radial Bar** | 🟢 Ranking | Presentations, cyclical data | 12-18 |

## Key Takeaways

::: {style="font-size: 0.85em;"}
::: {.incremental}
**When to use each:**
- **Lollipop > Bar:** When you have many categories or want cleaner look
- **Cleveland Dot:** Perfect for before/after or male/female comparisons  
- **Mosaic:** Test if two categorical variables are independent
- **Radial:** Use sparingly (harder to read), good for presentations

**Pro Tips:**
- `fct_reorder()` from forcats package is essential for proper ordering
- Mosaic aligned tiles = independent, misaligned = associated
- Remove gridlines for cleaner lollipop charts
- Always label mosaic plot axes clearly (area = proportion!)

**Common Mistakes:**
- Using radial charts for data without circular nature
- Too many categories in mosaic plots (>12 becomes unreadable)
- Not ordering lollipop charts by value
- Misinterpreting mosaic plots (height alone doesn't show proportion)
:::
:::

## Practice Exercise

**Your Turn:**

1. Create a lollipop chart of patient count by `ChronicDisease` type
2. Make a Cleveland dot plot comparing mean `SysBP_0m` by `SmokingStatus` and `Gender`
3. Create a mosaic plot testing independence of `AlcoholUse` and `SmokingStatus`
4. Interpret: Are alcohol use and smoking independent or associated?

## Next Steps {.center}

**Categorical Intermediate Complete!**

**Next:** Continuous Intermediate (Deck 2.2)

Learn advanced continuous visualizations:
- Violin plots
- Ridge plots
- 2D density
- Enhanced scatter plots
