---
title: "Continuous Plots - INTERMEDIATE"
subtitle: "Advanced Statistical Visualizations"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Continuous Intermediate {.center}

## Enhanced Continuous Visualizations

::: {style="font-size: 0.9em;"}
**Building on Basic Skills:**

Moving beyond simple histograms and scatter plots to sophisticated statistical visualizations.

**New Techniques:**
- Violin plots (distribution + density)
- Ridgeline plots (comparing many distributions)
- 2D density plots (large scatter data)
- Hexbin plots (overplotting solution)
- Scatter with regression (trend lines)

**New Packages:**
- `ggridges` for ridgeline plots
- `hexbin` for hexagonal binning
- Statistical layers in ggplot2
:::

## Setup

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-5|7-8|10-12"

# Load required libraries
library(ggplot2)
library(dplyr)
library(ggridges)    # For ridgeline plots
library(hexbin)      # For hexbin plots

# Load dataset
data <- readxl::read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Clean data: complete cases for analysis
data_clean <- data %>%
  filter(!is.na(BMI), !is.na(Cholesterol), !is.na(ExerciseFreq))
```

## Data Preview {.smaller}

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(ggridges)
library(hexbin)
library(readxl)

data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")
data_clean <- data %>% filter(!is.na(BMI), !is.na(Cholesterol), !is.na(ExerciseFreq))

data_clean %>%
  group_by(Region) %>%
  summarise(
    Count = n(),
    Mean_BMI = round(mean(BMI), 2),
    SD_BMI = round(sd(BMI), 2),
    Mean_Chol = round(mean(Cholesterol), 1)
  )
```

---

# Violin Plot

## Violin Plot - Code

🟡 **Goal:** Show full distribution shape by group

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3-4|5-7|8-10"

# Create violin plot comparing BMI across regions
ggplot(data_clean, aes(x = Region, y = BMI, fill = Region)) +
  geom_violin(alpha = 0.7, trim = FALSE) +  # Create violin shapes
  geom_boxplot(width = 0.15, alpha = 0.5, outlier.shape = NA) +  # Add boxplot inside
  scale_fill_brewer(palette = "Set2", name = "Region") +  # Color palette
  labs(title = "BMI Distribution by Region (Violin Plot)",
       x = "Region", y = "Body Mass Index (BMI)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        legend.position = "none")  # X-axis labels regions
```

::: {.notes}
Violin plot = rotated density plot on each side. Wider = more data at that value
:::

## Violin Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = Region, y = BMI, fill = Region)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.15, alpha = 0.5, outlier.shape = NA) +
  scale_fill_brewer(palette = "Set2", name = "Region") +
  labs(title = "BMI Distribution by Region (Violin Plot)",
       x = "Region", y = "Body Mass Index (BMI)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")
```

---

# Interpreting Violin Plots

## How to Read Violin Plots

**Using the previous plot:**

::: {.incremental style="font-size: 0.85em;"}
**OBSERVATION:**
- All regions show symmetric, roughly normal distributions
- Widest parts (bulges) around BMI 25 for most regions
- South region: slightly wider violin (more variation)
- Boxplots inside show median and quartiles
- No extreme outliers visible

**INTERPRETATION:**
- BMI distributions similar across all regions
- Most patients cluster around BMI 25 (normal/overweight boundary)
- South region has more variability in BMI
- No regional bias in BMI recruitment
- Distributions roughly normal (good for parametric tests)

**IMPLICATIONS:**
- Regional effects on outcomes unlikely due to BMI differences
- Can pool regions for BMI-related analyses
- South's higher variability may need investigation
- Sample representative across regions for BMI

**ACTIONS:**
- Proceed with region-pooled analyses for BMI
- Consider South region variability in sensitivity analyses
- Use BMI as continuous covariate (not categorized by region)
:::
:::

---

# Ridgeline Plot

## Ridgeline Plot - Code

🟡 **Goal:** Compare many distributions elegantly

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-3|5-7|8-11|12-14"

# Create ridgeline plot of Cholesterol by Exercise Frequency
# (overlapping density plots, like mountain ridges)
ggplot(data_clean, aes(x = Cholesterol, y = ExerciseFreq, fill = ExerciseFreq)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +  # Create overlapping densities
  scale_fill_brewer(palette = "Set2", name = "Exercise Frequency") +
  labs(title = "Cholesterol Distribution by Exercise Frequency",
       x = "Cholesterol (mg/dL)", 
       y = "Exercise Frequency") +
  theme_ridges(font_family = "serif") +  # Special theme for ridgeline plots
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        legend.position = "none")
```

::: {.notes}
Ridgeline plots are perfect when comparing 3+ groups. Scale parameter controls overlap
:::

## Ridgeline Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false
#| warning: false

ggplot(data_clean, aes(x = Cholesterol, y = ExerciseFreq, fill = ExerciseFreq)) +
  geom_density_ridges(alpha = 0.7, scale = 1.2) +
  scale_fill_brewer(palette = "Set2", name = "Exercise Frequency") +
  labs(title = "Cholesterol Distribution by Exercise Frequency",
       x = "Cholesterol (mg/dL)", 
       y = "Exercise Frequency") +
  theme_ridges(font_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")
```

---

# 2D Density Plot

## 2D Density Plot - Code

⚫ **Goal:** Show correlation patterns with density contours

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3-4|5-7|8-10"

# Create 2D density contour plot for Age vs. Cholesterol
ggplot(data_clean, aes(x = Age, y = Cholesterol)) +
  geom_density_2d_filled(alpha = 0.8, contour_var = "ndensity") +  # Filled contours
  geom_point(size = 0.5, alpha = 0.3, color = "white") +  # Add semi-transparent points
  scale_fill_viridis_d(option = "plasma", name = "Density") +  # Color scale
  labs(title = "2D Density: Age vs. Cholesterol Relationship",
       x = "Age (years)", 
       y = "Cholesterol (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12))
```

::: {.notes}
2D density shows where most data points cluster. Darker = higher concentration
:::

## 2D Density Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = Age, y = Cholesterol)) +
  geom_density_2d_filled(alpha = 0.8, contour_var = "ndensity") +
  geom_point(size = 0.5, alpha = 0.3, color = "white") +
  scale_fill_viridis_d(option = "plasma", name = "Density") +
  labs(title = "2D Density: Age vs. Cholesterol Relationship",
       x = "Age (years)", 
       y = "Cholesterol (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"))
```

---

# Hexbin Plot

## Hexbin Plot - Code

⚫ **Goal:** Handle overplotting in large scatter plots

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3-4|5-7|8-10"

# Create hexbin plot for BMI vs. Blood Sugar (handles overlapping points)
ggplot(data_clean, aes(x = BMI, y = BloodSugar)) +
  geom_hex(bins = 30) +  # Create hexagonal bins
  scale_fill_viridis_c(option = "inferno", name = "Count") +  # Color by count
  labs(title = "BMI vs. Blood Sugar (Hexbin Plot)",
       x = "Body Mass Index (BMI)", 
       y = "Blood Sugar (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12))
```

::: {.notes}
Hexbin divides plot into hexagons and counts points in each. Better than overlapping points
:::

## Hexbin Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = BMI, y = BloodSugar)) +
  geom_hex(bins = 30) +
  scale_fill_viridis_c(option = "inferno", name = "Count") +
  labs(title = "BMI vs. Blood Sugar (Hexbin Plot)",
       x = "Body Mass Index (BMI)", 
       y = "Blood Sugar (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"))
```

---

# Scatter with Regression

## Scatter with Regression Line - Code

⚫ **Goal:** Show correlation with statistical trend

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3-4|5-8|9-13|14-16"

# Scatter plot with linear regression line and confidence interval
ggplot(data_clean, aes(x = Age, y = SysBP_0m)) +
  geom_point(alpha = 0.5, color = "steelblue", size = 2.5) +  # Add points
  geom_smooth(method = "lm", se = TRUE, color = "#E74C3C", 
              fill = "#E74C3C", alpha = 0.2, size = 1.2) +  # Add regression line + CI
  # Calculate and display correlation
  annotate("text", x = 25, y = 160, 
           label = paste("r =", round(cor(data_clean$Age, data_clean$SysBP_0m), 3)),
           family = "serif", size = 5, fontface = "bold") +
  labs(title = "Age vs. Baseline Systolic BP with Regression Line",
       subtitle = "Linear trend with 95% confidence interval",
       x = "Age (years)", 
       y = "Systolic Blood Pressure (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        axis.title = element_text(size = 12))
```

## Scatter with Regression - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = Age, y = SysBP_0m)) +
  geom_point(alpha = 0.5, color = "steelblue", size = 2.5) +
  geom_smooth(method = "lm", se = TRUE, color = "#E74C3C", 
              fill = "#E74C3C", alpha = 0.2, size = 1.2) +
  annotate("text", x = 25, y = 160, 
           label = paste("r =", round(cor(data_clean$Age, data_clean$SysBP_0m), 3)),
           family = "serif", size = 5, fontface = "bold") +
  labs(title = "Age vs. Baseline Systolic BP with Regression Line",
       subtitle = "Linear trend with 95% confidence interval",
       x = "Age (years)", 
       y = "Systolic Blood Pressure (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5, size = 12),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

---

# Interpreting Density & Regression

## Reading Advanced Continuous Plots

**Using the regression plot:**

::: {.incremental style="font-size: 0.85em;"}
**OBSERVATION:**
- Weak positive correlation (r ≈ 0.2-0.3)
- Regression line slopes upward slightly
- Wide confidence interval (uncertainty in relationship)
- Substantial scatter around the line
- Age range: 18-79 years, BP range: 100-165 mmHg

**INTERPRETATION:**
- Age explains only small portion of BP variation (~4-9% via r²)
- Systolic BP tends to increase slightly with age (expected)
- Many other factors influence BP beyond age
- Relationship is linear but weak
- Individual variation is large

**IMPLICATIONS:**
- Age alone is poor predictor of BP
- Need multivariate models (BMI, smoking, etc.)
- Age adjustment important but not sufficient
- Biological relationship exists but is noisy
- Clinical: Don't assume high BP from age alone

**CRITICAL REMINDERS:**
- **Correlation ≠ Causation** (age doesn't "cause" high BP directly)
- r < 0.3 = weak, 0.3-0.7 = moderate, >0.7 = strong
- Always check confidence interval width
- Look for non-linear patterns
:::
:::

---

# Quick Reference

## Continuous Intermediate - Summary

| Chart Type | Goal | When to Use | Code Lines |
|------------|------|-------------|-----------|
| **Violin Plot** | 🟡 Distribution | Compare 2-6 groups with full distribution | 10-15 |
| **Ridgeline** | 🟡 Distribution | Compare 3+ groups elegantly | 12-18 |
| **2D Density** | ⚫ Correlation | Show clustering in scatter data | 10-15 |
| **Hexbin** | ⚫ Correlation | Handle overplotting (1000+ points) | 10-15 |
| **Scatter + Regression** | ⚫ Correlation | Show trend + uncertainty | 15-20 |

## Violin vs. Box Plot

::: {.columns style="font-size: 0.85em;"}
::: {.column width="50%"}
**Box Plot Shows:**
- 5-number summary
- Median (50th percentile)
- Quartiles (25th, 75th)
- Outliers
- **Missing:** Distribution shape

**Use When:**
- Quick comparison needed
- Focus on median/quartiles
- Outliers are important
:::

::: {.column width="50%"}
**Violin Plot Shows:**
- Full distribution shape
- Density at each value
- Can see bimodal patterns
- Shows skewness clearly
- **Missing:** Exact quantiles

**Use When:**
- Shape matters
- Want to see multiple peaks
- Comparing distributions
- Publication quality needed
:::
:::

## Key Takeaways

::: {style="font-size: 0.85em;"}
::: {.incremental}
**Remember:**

1. **Violin plots** show what box plots hide (distribution shape)
2. **Ridgeline plots** scale well to 5-10 groups
3. **2D density** reveals clustering patterns
4. **Hexbin** solves overplotting automatically
5. **Always show confidence intervals** on regression lines

**Common Mistakes:**
- Using violin plots with < 20 observations per group
- Too many groups in ridgeline (>10 gets cluttered)
- Forgetting to show confidence intervals
- Interpreting weak correlations as strong relationships
- Not checking for non-linear patterns

**Pro Tips:**
- Add boxplot inside violin for best of both worlds
- `scale` parameter in ridgeline controls overlap
- Use `alpha` transparency to show density overlaps
- Always report correlation coefficient with regression
:::
:::

## Practice Exercise

**Your Turn:**

1. Create a violin plot of `Cholesterol` by `SmokingStatus`
2. Add a small boxplot inside each violin
3. Create a ridgeline plot of `Age` distribution by `Region`
4. Make a scatter plot with regression line: `BMI` vs. `Cholesterol`
5. Add the correlation coefficient as text annotation

**Bonus:** Create a hexbin plot colored by density for `Age` vs. `BloodSugar`

## Next Steps {.center}

**Continuous Intermediate Complete!**

**Next:** Category vs. Continuous Intermediate (Deck 2.3)

Learn multi-dimensional visualizations:
- Treemaps
- Heatmaps
- Faceted plots
- 100% stacked bars
- Slope charts
