---
title: "Category vs. Continuous - INTERMEDIATE"
subtitle: "Advanced Group Comparisons"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Category vs. Continuous {.center}

## What We'll Learn - Intermediate

::: {.columns}
::: {.column width="50%"}
**Beyond Basic Box Plots:**

- Violin plots (distribution shape)
- Ridgeline plots (multiple distributions)
- Bee swarm plots (all data points)
- Raincloud plots (comprehensive view)
- Enhanced grouped/stacked bars
:::

::: {.column width="50%"}
**Analytical Goals:**

- 🟡 **Distribution** - See full shapes
- 🟢 **Ranking** - Compare multiple groups
- ⚫ **Correlation** - Patterns across categories
- 🔴 **Part of Whole** - Compositional analysis
:::
:::

## Setup Code

```{r}
#| echo: true
#| message: false

# Load required packages
library(readxl)
library(dplyr)
library(ggplot2)
library(ggridges)  # For ridgeline plots
library(ggbeeswarm) # For bee swarm plots

# Load dataset
health_data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Create custom theme
theme_custom <- function() {
  theme_minimal() +
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.title = element_text(size = 11),
    legend.position = "bottom"
  )
}
```

## Violin Plot - Code

::: {style="font-size: 0.85em;"}
**Purpose:** Show full distribution shape within each category

**Goal:** 🟡 Distribution

```{r}
#| echo: true
#| eval: false

# Create violin plot: BMI distribution by Gender
ggplot(health_data, aes(x = Gender, y = BMI, fill = Gender)) +
  geom_violin(trim = FALSE,        # Don't trim tails
              alpha = 0.7) +       # Transparency
  geom_boxplot(width = 0.1,       # Add boxplot inside
               fill = "white",
               outlier.shape = NA) + # Hide outliers (shown in violin)
  stat_summary(fun = mean,         # Add mean point
               geom = "point",
               shape = 18,
               size = 3,
               color = "red") +
  labs(title = "BMI Distribution by Gender",
       subtitle = "Violin plot shows full distribution shape",
       x = "Gender",
       y = "Body Mass Index (kg/m²)") +
  scale_fill_manual(values = c("Male" = "#4DBBD5", "Female" = "#E64B35")) +
  theme_custom() +
  theme(legend.position = "none")  # Remove legend (redundant with x-axis)
```
:::

## Violin Plot - Output

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 5

# Create violin plot
ggplot(health_data, aes(x = Gender, y = BMI, fill = Gender)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  labs(title = "BMI Distribution by Gender",
       subtitle = "Violin plot shows full distribution shape",
       x = "Gender",
       y = "Body Mass Index (kg/m²)") +
  scale_fill_manual(values = c("Male" = "#4DBBD5", "Female" = "#E64B35")) +
  theme_custom() +
  theme(legend.position = "none")
```

## Interpreting Violin Plots

::: {.columns}
::: {.column width="50%"}
**What to Look For:**

- **Width** = Density of data
- **Wider sections** = More observations
- **Narrower sections** = Fewer observations
- **Shape** reveals distribution pattern

**Compare shapes:**

- Symmetry vs. skewness
- Single peak vs. multiple peaks
- Spread differences
:::

::: {.column width="50%"}
**In Our Example:**

- Both genders show similar BMI distributions
- Slight right skew (tail toward higher BMI)
- Males show slightly wider spread
- Boxplot inside shows quartiles
- Red diamond = mean
- Most data concentrated in 25-30 range
:::
:::

## Ridgeline Plot - Code

::: {style="font-size: 0.85em;"}
**Purpose:** Compare distributions across many categories

**Goal:** 🟡 Distribution + 🟢 Ranking

```{r}
#| echo: true
#| eval: false

# Create ridgeline plot: Age distribution by Exercise Frequency
health_data %>%
  mutate(ExerciseFreq = factor(ExerciseFreq,
                              levels = c("5-7 days/week", "3-4 days/week",
                                       "1-2 days/week", "Never"))) %>%
  ggplot(aes(x = Age, y = ExerciseFreq, fill = ExerciseFreq)) +
  geom_density_ridges(alpha = 0.7,        # Transparency
                     scale = 1.5,        # Overlap amount
                     rel_min_height = 0.01) + # Trim small tails
  labs(title = "Age Distribution by Exercise Frequency",
       subtitle = "Ridgeline plot shows overlapping distributions",
       x = "Age (years)",
       y = "Exercise Frequency") +
  scale_fill_viridis_d(option = "plasma") +  # Color-blind friendly
  theme_custom() +
  theme(legend.position = "none")
```
:::

## Ridgeline Plot - Output

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6

library(ggridges)

health_data %>%
  mutate(ExerciseFreq = factor(ExerciseFreq,
                              levels = c("5-7 days/week", "3-4 days/week",
                                       "1-2 days/week", "Never"))) %>%
  ggplot(aes(x = Age, y = ExerciseFreq, fill = ExerciseFreq)) +
  geom_density_ridges(alpha = 0.7, scale = 1.5, rel_min_height = 0.01) +
  labs(title = "Age Distribution by Exercise Frequency",
       subtitle = "Ridgeline plot shows overlapping distributions",
       x = "Age (years)",
       y = "Exercise Frequency") +
  scale_fill_viridis_d(option = "plasma") +
  theme_custom() +
  theme(legend.position = "none")
```

## Interpreting Ridgeline Plots

**Key Observations:**

::: {.incremental}
1. **People who never exercise tend to be older**
   - Peak around 65-70 years

2. **More frequent exercisers tend to be younger**
   - "5-7 days/week" peaks around 35-40 years

3. **All groups show wide age ranges**
   - Exercise happens at all ages
   - But frequency patterns differ

4. **The overlap shows transitions**
   - Exercise habits change across life stages
:::

**When to use:** Comparing 3+ distributions, especially when order matters

## Bee Swarm Plot - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show every individual data point without overlap

**Goal:** 🟡 Distribution + individual observations

```{r}
#| echo: true
#| eval: false

# Create bee swarm plot: Cholesterol by Smoking Status
health_data %>%
  filter(!is.na(SmokingStatus), !is.na(Cholesterol)) %>%
  mutate(SmokingStatus = factor(SmokingStatus,
                               levels = c("Never", "Former", "Current"))) %>%
  ggplot(aes(x = SmokingStatus, y = Cholesterol, color = SmokingStatus)) +
  geom_quasirandom(alpha = 0.5,      # Transparency for overlaps
                  size = 2) +        # Point size
  stat_summary(fun = median,         # Add median line
               geom = "crossbar",
               width = 0.5,
               color = "black",
               linewidth = 0.8) +
  stat_summary(fun.data = mean_se,   # Add mean with error bars
               geom = "pointrange",
               color = "red",
               size = 0.8) +
  labs(title = "Cholesterol Levels by Smoking Status",
       subtitle = "Each point is one patient | Black bar = median | Red = mean ± SE",
       x = "Smoking Status",
       y = "Total Cholesterol (mg/dL)") +
  scale_color_manual(values = c("Never" = "#00A087", "Former" = "#3C5488", 
                               "Current" = "#DC0000")) +
  theme_custom() +
  theme(legend.position = "none")
```
:::

## Bee Swarm Plot - Output

```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 6
#| warning: false

library(ggbeeswarm)

health_data %>%
  filter(!is.na(SmokingStatus), !is.na(Cholesterol)) %>%
  mutate(SmokingStatus = factor(SmokingStatus,
                               levels = c("Never", "Former", "Current"))) %>%
  ggplot(aes(x = SmokingStatus, y = Cholesterol, color = SmokingStatus)) +
  geom_quasirandom(alpha = 0.5, size = 2) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, 
               color = "black", linewidth = 0.8) +
  stat_summary(fun.data = mean_se, geom = "pointrange", 
               color = "red", size = 0.8) +
  labs(title = "Cholesterol Levels by Smoking Status",
       subtitle = "Each point is one patient | Black bar = median | Red = mean ± SE",
       x = "Smoking Status",
       y = "Total Cholesterol (mg/dL)") +
  scale_color_manual(values = c("Never" = "#00A087", "Former" = "#3C5488", 
                               "Current" = "#DC0000")) +
  theme_custom() +
  theme(legend.position = "none")
```

## Interpreting Bee Swarm Plots

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**Advantages:**

- **See every data point** - no data hiding
- **Identify outliers** easily
- **Compare distributions** visually
- **Show sample sizes** (more points = more data)
- **Patterns emerge** from cloud shapes

**When to use:**

- Small to medium datasets (n < 500 per group)
- Want to show individual variation
- Need to identify specific outliers
:::

::: {.column width="50%"}
**Our Findings:**

1. **Current smokers** have slightly higher cholesterol
   - Median around 200 mg/dL

2. **Former smokers** show wide variation
   - Suggests health improvements vary by person

3. **Several high outliers** in all groups
   - These patients need attention

4. **Sample sizes visible**
   - "Never" has most patients
   - "Current" has fewest
:::
:::

## Raincloud Plot - Code

::: {style="font-size: 0.75em;"}
**Purpose:** Comprehensive view - combines violin + points + box

**Goal:** 🟡 Distribution (most complete visualization)

```{r}
#| echo: true
#| eval: false

# Create raincloud plot: Blood Pressure by Treatment Group
library(ggdist)  # For stat_halfeye

health_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP_6m)) %>%
  ggplot(aes(x = TreatmentGroup, y = SysBP_6m, fill = TreatmentGroup)) +
  # Half violin (cloud)
  stat_halfeye(adjust = 0.5, width = 0.6, .width = 0, 
               justification = -0.2, point_colour = NA) +
  # Box plot
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5,
               position = position_nudge(x = -0.1)) +
  # Individual points (rain)
  geom_point(aes(color = TreatmentGroup), size = 1.3, alpha = 0.3,
             position = position_jitter(seed = 1, width = 0.05,
                                       height = 0)) +
  # Add mean point
  stat_summary(fun = mean, geom = "point", shape = 18, 
               size = 4, color = "black") +
  coord_flip() +  # Flip to horizontal
  labs(title = "Systolic Blood Pressure at 6 Months by Treatment Group",
       subtitle = "Raincloud plot: density (cloud) + boxplot + points (rain)",
       x = "Treatment Group",
       y = "Systolic Blood Pressure (mmHg)") +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  theme_custom() +
  theme(legend.position = "none")
```
:::

## Raincloud Plot - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
#| message: false

library(ggdist)

health_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP_6m)) %>%
  ggplot(aes(x = TreatmentGroup, y = SysBP_6m, fill = TreatmentGroup)) +
  stat_halfeye(adjust = 0.5, width = 0.6, .width = 0, 
               justification = -0.2, point_colour = NA) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.5,
               position = position_nudge(x = -0.1)) +
  geom_point(aes(color = TreatmentGroup), size = 1.3, alpha = 0.3,
             position = position_jitter(seed = 1, width = 0.05, height = 0)) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "black") +
  coord_flip() +
  labs(title = "Systolic Blood Pressure at 6 Months by Treatment Group",
       subtitle = "Raincloud plot: density (cloud) + boxplot + points (rain)",
       x = "Treatment Group",
       y = "Systolic Blood Pressure (mmHg)") +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  theme_custom() +
  theme(legend.position = "none")
```

## Interpreting Raincloud Plots

::: {style="font-size: 0.85em;"}
**Three Components:**

1. **Cloud (density curve)** - Shows distribution shape
2. **Box plot** - Shows quartiles and median
3. **Rain (points)** - Shows actual data

**What We See:**

::: {.incremental}
- **Control group** has highest BP (expected)
- **Treatment groups** show reduced BP
- **All groups** show similar spread (IQR)
- **Some patients** don't respond well (upper outliers in treatment groups)
- **Distribution shapes** are roughly similar (slightly right-skewed)
- **Sample sizes** visible from number of points
:::

**Advantage:** Most informative single plot for comparing distributions
:::

## Enhanced Grouped Bar Chart - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Compare means across two categorical variables

**Goal:** 🟢 Ranking + ⚫ Correlation

```{r}
#| echo: true
#| eval: false

# Calculate summary statistics
summary_data <- health_data %>%
  filter(!is.na(Gender), !is.na(ExerciseFreq), !is.na(BMI)) %>%
  group_by(Gender, ExerciseFreq) %>%
  summarise(
    mean_bmi = mean(BMI, na.rm = TRUE),
    se_bmi = sd(BMI, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(ExerciseFreq = factor(ExerciseFreq,
                              levels = c("Never", "1-2 days/week",
                                       "3-4 days/week", "5-7 days/week")))

# Create grouped bar chart with error bars and sample sizes
ggplot(summary_data, aes(x = ExerciseFreq, y = mean_bmi, fill = Gender)) +
  geom_col(position = position_dodge(0.9), alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_bmi - se_bmi, ymax = mean_bmi + se_bmi),
                position = position_dodge(0.9), width = 0.25) +
  geom_text(aes(label = paste0("n=", n)), 
            position = position_dodge(0.9), vjust = -1.5, size = 3) +
  labs(title = "Mean BMI by Exercise Frequency and Gender",
       subtitle = "Error bars show ± 1 SE | Numbers show sample size",
       x = "Exercise Frequency",
       y = "Mean BMI (kg/m²)",
       fill = "Gender") +
  scale_fill_manual(values = c("Male" = "#4DBBD5", "Female" = "#E64B35")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_custom()
```
:::

## Enhanced Grouped Bar Chart - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false

summary_data <- health_data %>%
  filter(!is.na(Gender), !is.na(ExerciseFreq), !is.na(BMI)) %>%
  group_by(Gender, ExerciseFreq) %>%
  summarise(
    mean_bmi = mean(BMI, na.rm = TRUE),
    se_bmi = sd(BMI, na.rm = TRUE) / sqrt(n()),
    n = n(),
    .groups = "drop"
  ) %>%
  mutate(ExerciseFreq = factor(ExerciseFreq,
                              levels = c("Never", "1-2 days/week",
                                       "3-4 days/week", "5-7 days/week")))

ggplot(summary_data, aes(x = ExerciseFreq, y = mean_bmi, fill = Gender)) +
  geom_col(position = position_dodge(0.9), alpha = 0.8) +
  geom_errorbar(aes(ymin = mean_bmi - se_bmi, ymax = mean_bmi + se_bmi),
                position = position_dodge(0.9), width = 0.25) +
  geom_text(aes(label = paste0("n=", n)), 
            position = position_dodge(0.9), vjust = -1.5, size = 3) +
  labs(title = "Mean BMI by Exercise Frequency and Gender",
       subtitle = "Error bars show ± 1 SE | Numbers show sample size",
       x = "Exercise Frequency",
       y = "Mean BMI (kg/m²)",
       fill = "Gender") +
  scale_fill_manual(values = c("Male" = "#4DBBD5", "Female" = "#E64B35")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  theme_custom()
```

## Interpreting Enhanced Grouped Bars

::: {style="font-size: 0.9em;"}
**Key Observations:**

::: {.incremental}
1. **Clear inverse relationship** - More exercise → Lower BMI
   - Consistent in both genders

2. **Gender differences** are small but consistent
   - Males slightly higher BMI across all groups
   - Error bars overlap → differences may not be significant

3. **Sample sizes matter**
   - "Never" group has largest n
   - "5-7 days/week" has smallest n
   - Affects confidence in estimates

4. **Error bars show uncertainty**
   - Wider bars = more variability
   - Consider both mean AND uncertainty when interpreting
:::
:::

## Faceted Box Plots - Code

::: {style="font-size: 0.85em;"}
**Purpose:** Compare distributions across multiple dimensions

**Goal:** 🟡 Distribution + ⚫ Patterns

```{r}
#| echo: true
#| eval: false

# Create faceted box plots: Cholesterol by Region and Gender
health_data %>%
  filter(!is.na(Region), !is.na(Gender), !is.na(Cholesterol)) %>%
  ggplot(aes(x = Gender, y = Cholesterol, fill = Gender)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, 
               size = 3, color = "red") +
  facet_wrap(~ Region, ncol = 2) +  # Split by Region
  labs(title = "Cholesterol Levels by Gender Across Regions",
       subtitle = "Faceted box plots reveal regional patterns | Red diamond = mean",
       x = "Gender",
       y = "Total Cholesterol (mg/dL)") +
  scale_fill_manual(values = c("Male" = "#4DBBD5", "Female" = "#E64B35")) +
  theme_custom() +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "gray90", color = "black"),
        strip.text = element_text(face = "bold", size = 11))
```
:::

## Faceted Box Plots - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7

health_data %>%
  filter(!is.na(Region), !is.na(Gender), !is.na(Cholesterol)) %>%
  ggplot(aes(x = Gender, y = Cholesterol, fill = Gender)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
  stat_summary(fun = mean, geom = "point", shape = 18, 
               size = 3, color = "red") +
  facet_wrap(~ Region, ncol = 2) +
  labs(title = "Cholesterol Levels by Gender Across Regions",
       subtitle = "Faceted box plots reveal regional patterns | Red diamond = mean",
       x = "Gender",
       y = "Total Cholesterol (mg/dL)") +
  scale_fill_manual(values = c("Male" = "#4DBBD5", "Female" = "#E64B35")) +
  theme_custom() +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "gray90", color = "black"),
        strip.text = element_text(face = "bold", size = 11))
```

## Interpreting Faceted Box Plots

::: {style="font-size: 0.9em;"}
**Regional Patterns:**

::: {.incremental}
1. **East region** shows highest cholesterol overall
   - Both genders affected

2. **West region** shows lowest levels
   - Possible dietary or lifestyle differences?

3. **Gender differences consistent** across all regions
   - Males tend to have slightly higher values

4. **Outliers present** in all regions
   - Need follow-up investigation

5. **Regional variation > Gender variation**
   - Geography matters more than gender for cholesterol
:::

**When to use facets:** When you have 3+ categorical variables to compare
:::

## Chart Selection Guide

::: {style="font-size: 0.85em;"}
| Chart Type | Goal | Best When | Sample Size |
|------------|------|-----------|-------------|
| **Violin Plot** | 🟡 Distribution shape | 2-5 groups, want to see shape | 50+ per group |
| **Ridgeline Plot** | 🟡 Distribution + 🟢 Ranking | 3+ groups, order matters | 30+ per group |
| **Bee Swarm** | 🟡 + individual points | Show every observation | < 500 per group |
| **Raincloud** | 🟡 Comprehensive view | Publication-quality figure | 50+ per group |
| **Grouped Bar** | 🟢 Ranking + ⚫ Comparison | Two categorical variables | Any size |
| **Faceted Box** | 🟡 Distribution + patterns | 3+ categorical variables | 20+ per group |

**Pro Tip:** Start with violin or raincloud for exploration, then choose based on audience needs
:::

## Common Mistakes to Avoid

::: {.incremental style="font-size: 0.9em;"}
1. **Too many groups** in one plot
   - Solution: Use facets or filter to top groups

2. **Ignoring sample size differences**
   - Solution: Show n on plot or use error bars

3. **Comparing distributions without uncertainty**
   - Solution: Add confidence intervals or error bars

4. **Bee swarms with too many points**
   - Solution: Use violin or density if n > 500

5. **Not ordering categories logically**
   - Solution: Order by median, mean, or natural order

6. **Forgetting about color-blind audiences**
   - Solution: Use color-blind safe palettes (viridis, ColorBrewer)
:::

## Missing Data Handling

::: {style="font-size: 0.9em;"}
**Always check for missing data:**

```{r}
#| echo: true
#| eval: false

# Check missing data by group
health_data %>%
  group_by(Gender, ExerciseFreq) %>%
  summarise(
    total = n(),
    missing_bmi = sum(is.na(BMI)),
    pct_missing = round(100 * missing_bmi / total, 1)
  )
```

**Options for handling:**

1. **Filter complete cases** - `filter(!is.na(variable))`
2. **Show as separate category** - useful if missing is informative
3. **Note in subtitle** - "Excludes N missing observations"
4. **Investigate patterns** - Is missingness related to outcome?
:::

## Practice Exercise

**Your Turn:**

1. Create a violin plot of `BloodSugar_6m` by `TreatmentGroup`
2. Add a ridgeline plot showing `Age` distribution by `Region`
3. Create a bee swarm plot of `Cholesterol_6m` by `SmokingStatus`
4. Build a raincloud plot of `SysBP_6m` by `Gender`

**Bonus Challenge:**

Create a faceted violin plot showing `BMI` by `Gender` across all 4 `Regions`, ordered by median BMI.

## Summary

::: {style="font-size: 0.9em;"}
**What You've Learned:**

::: {.incremental}
1. **Violin plots** show full distribution shapes
2. **Ridgeline plots** compare many overlapping distributions
3. **Bee swarm plots** show every data point
4. **Raincloud plots** provide comprehensive views
5. **Enhanced grouped bars** show means with uncertainty
6. **Faceted plots** reveal multi-dimensional patterns

**Key Principle:**

> Choose charts that match your data size, analytical goal, and audience needs
:::
:::

## Next Steps {.center}

**You've mastered Category vs. Continuous - Intermediate!**

**Next:** Time Series - Intermediate

Learn advanced time series visualization with:

- Smoothed trends
- Seasonal decomposition
- Multiple series comparisons
- Annotations and forecasts
