---
title: "Time Series - INTERMEDIATE"
subtitle: "Advanced Temporal Analysis"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Time Series Analysis {.center}

## What We'll Learn - Intermediate

::: {.columns}
::: {.column width="50%"}
**Beyond Simple Line Charts:**

- Smoothed trends (LOESS, moving average)
- Area charts with confidence bands
- Multi-series comparison plots
- Annotations and event markers
- Seasonal patterns
- Change highlighting
:::

::: {.column width="50%"}
**Analytical Goals:**

- 🔵 **Evolution** - How things change over time
- 🟡 **Distribution** - Spread at different times
- 🟢 **Ranking** - Compare multiple series
- ⚫ **Correlation** - Relationship between series
:::
:::

## Setup Code

```{r}
#| echo: true
#| message: false

# Load required packages
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(zoo)      # For moving averages
library(lubridate) # For date handling

# Load dataset and create time series data
health_data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Aggregate by time point
time_data <- health_data %>%
  pivot_longer(cols = c(SysBP_0m, SysBP_3m, SysBP_6m),
               names_to = "TimePoint",
               values_to = "SysBP") %>%
  mutate(Month = case_when(
    TimePoint == "SysBP_0m" ~ 0,
    TimePoint == "SysBP_3m" ~ 3,
    TimePoint == "SysBP_6m" ~ 6
  ))

# Create theme
theme_custom <- function() {
  theme_minimal() +
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.title = element_text(size = 11),
    legend.position = "bottom"
  )
}
```

## Smoothed Trend with LOESS - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show underlying trend while reducing noise

**Goal:** 🔵 Evolution

```{r}
#| echo: true
#| eval: false

# Calculate mean BP by treatment group over time
trend_data <- time_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP)) %>%
  group_by(TreatmentGroup, Month) %>%
  summarise(
    mean_bp = mean(SysBP, na.rm = TRUE),
    se_bp = sd(SysBP, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Create smoothed trend plot
ggplot(trend_data, aes(x = Month, y = mean_bp, 
                      color = TreatmentGroup, fill = TreatmentGroup)) +
  # Raw data points
  geom_point(size = 3, alpha = 0.6) +
  # Smoothed trend line
  geom_smooth(method = "loess",     # Locally weighted smoothing
              span = 0.75,          # Smoothness (larger = smoother)
              alpha = 0.2) +        # Confidence band transparency
  labs(title = "Blood Pressure Trends by Treatment Group",
       subtitle = "Smoothed LOESS curves with 95% confidence bands",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)",
       color = "Treatment",
       fill = "Treatment") +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom()
```
:::

## Smoothed Trend with LOESS - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| warning: false
#| message: false

trend_data <- time_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP)) %>%
  group_by(TreatmentGroup, Month) %>%
  summarise(
    mean_bp = mean(SysBP, na.rm = TRUE),
    se_bp = sd(SysBP, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(trend_data, aes(x = Month, y = mean_bp, 
                      color = TreatmentGroup, fill = TreatmentGroup)) +
  geom_point(size = 3, alpha = 0.6) +
  geom_smooth(method = "loess", span = 0.75, alpha = 0.2) +
  labs(title = "Blood Pressure Trends by Treatment Group",
       subtitle = "Smoothed LOESS curves with 95% confidence bands",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)",
       color = "Treatment",
       fill = "Treatment") +
  scale_color_brewer(palette = "Set1") +
  scale_fill_brewer(palette = "Set1") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom()
```

## Interpreting Smoothed Trends

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**LOESS Smoothing:**

- **Local regression** at each point
- Reduces random fluctuations
- Reveals underlying patterns
- Confidence bands show uncertainty

**Span parameter:**

- `span = 0.3` - Very wiggly (follows data closely)
- `span = 0.75` - Moderate smoothing (default)
- `span = 1.5` - Very smooth (may miss details)
:::

::: {.column width="50%"}
**Our Findings:**

1. **All groups start similar** at Month 0

2. **Treatment A & B effective**
   - Clear downward trend
   - BP drops ~15-20 mmHg

3. **Control group stable**
   - No significant change
   - As expected

4. **Confidence bands widen** over time
   - Less certainty in estimates
   - Possible dropout effects
:::
:::

## Area Chart with Confidence Bands - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Emphasize magnitude and uncertainty over time

**Goal:** 🔵 Evolution + 🟡 Distribution

```{r}
#| echo: true
#| eval: false

# Calculate confidence intervals manually
trend_ci <- trend_data %>%
  mutate(
    lower = mean_bp - 1.96 * se_bp,  # 95% CI
    upper = mean_bp + 1.96 * se_bp
  )

# Create area chart with confidence band
ggplot(trend_ci, aes(x = Month, y = mean_bp, 
                    color = TreatmentGroup, fill = TreatmentGroup)) +
  # Confidence band as ribbon
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              alpha = 0.3, color = NA) +
  # Mean line
  geom_line(linewidth = 1.2) +
  # Points at measurement times
  geom_point(size = 3) +
  labs(title = "Blood Pressure Changes with 95% Confidence Intervals",
       subtitle = "Shaded areas show uncertainty in estimates",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)",
       color = "Treatment",
       fill = "Treatment") +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom()
```
:::

## Area Chart with Confidence Bands - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

trend_ci <- trend_data %>%
  mutate(
    lower = mean_bp - 1.96 * se_bp,
    upper = mean_bp + 1.96 * se_bp
  )

ggplot(trend_ci, aes(x = Month, y = mean_bp, 
                    color = TreatmentGroup, fill = TreatmentGroup)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), 
              alpha = 0.3, color = NA) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(title = "Blood Pressure Changes with 95% Confidence Intervals",
       subtitle = "Shaded areas show uncertainty in estimates",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)",
       color = "Treatment",
       fill = "Treatment") +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom()
```

## Interpreting Area Charts

::: {style="font-size: 0.9em;"}
**Key Observations:**

::: {.incremental}
1. **Non-overlapping bands** = Statistically different
   - Treatment A vs Control clearly different at 6 & 12 months

2. **Overlapping bands** = May not be different
   - Treatment A vs B overlap at all time points
   - Suggests similar effectiveness

3. **Band width** shows precision
   - Narrower = more precise estimate
   - Wider = more uncertainty

4. **Trend direction** immediately visible
   - Downward = improvement
   - Flat = no change
   - Upward = worsening
:::
:::

## Annotated Time Series - Code

::: {style="font-size: 0.75em;"}
**Purpose:** Highlight important events and changes

**Goal:** 🔵 Evolution + context

```{r}
#| echo: true
#| eval: false

# Prepare data for one treatment group
treatment_a_data <- trend_ci %>% filter(TreatmentGroup == "Treatment A")

# Create annotated plot
ggplot(treatment_a_data, aes(x = Month, y = mean_bp)) +
  # Area under curve
  geom_area(fill = "#4DBBD5", alpha = 0.3) +
  # Line and points
  geom_line(color = "#4DBBD5", linewidth = 1.5) +
  geom_point(size = 4, color = "#4DBBD5") +
  # Annotate starting point
  annotate("text", x = 0, y = 142, 
           label = "Baseline\n140 mmHg", 
           hjust = 0.5, vjust = -0.5, size = 4, fontface = "bold") +
  # Annotate 6-month check
  annotate("segment", x = 6, y = 125, xend = 6, yend = 130,
           arrow = arrow(length = unit(0.3, "cm")), 
           color = "red", linewidth = 1) +
  annotate("text", x = 6, y = 123, 
           label = "First improvement\n-15 mmHg", 
           hjust = 0.5, size = 3.5, color = "red") +
  # Annotate final result
  annotate("text", x = 12, y = 118, 
           label = "Final: 120 mmHg\n✓ Target achieved", 
           hjust = 0.5, vjust = 1.2, size = 4, fontface = "bold", color = "darkgreen") +
  # Add horizontal reference line for target BP
  geom_hline(yintercept = 120, linetype = "dashed", 
             color = "darkgreen", linewidth = 0.8) +
  labs(title = "Treatment A: Blood Pressure Reduction Over 12 Months",
       subtitle = "Green line shows target BP of 120 mmHg",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  ylim(110, 145) +
  theme_custom()
```
:::

## Annotated Time Series - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

treatment_a_data <- trend_ci %>% filter(TreatmentGroup == "Treatment A")

ggplot(treatment_a_data, aes(x = Month, y = mean_bp)) +
  geom_area(fill = "#4DBBD5", alpha = 0.3) +
  geom_line(color = "#4DBBD5", linewidth = 1.5) +
  geom_point(size = 4, color = "#4DBBD5") +
  annotate("text", x = 0, y = 142, 
           label = "Baseline\n140 mmHg", 
           hjust = 0.5, vjust = -0.5, size = 4, fontface = "bold") +
  annotate("segment", x = 6, y = 125, xend = 6, yend = 130,
           arrow = arrow(length = unit(0.3, "cm")), 
           color = "red", linewidth = 1) +
  annotate("text", x = 6, y = 123, 
           label = "First improvement\n-15 mmHg", 
           hjust = 0.5, size = 3.5, color = "red") +
  annotate("text", x = 12, y = 118, 
           label = "Final: 120 mmHg\n✓ Target achieved", 
           hjust = 0.5, vjust = 1.2, size = 4, fontface = "bold", color = "darkgreen") +
  geom_hline(yintercept = 120, linetype = "dashed", 
             color = "darkgreen", linewidth = 0.8) +
  labs(title = "Treatment A: Blood Pressure Reduction Over 12 Months",
       subtitle = "Green line shows target BP of 120 mmHg",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  ylim(110, 145) +
  theme_custom()
```

## Effective Annotations

::: {style="font-size: 0.9em;"}
**When to Annotate:**

::: {.incremental}
- **Key milestones** - Important time points
- **Threshold crossings** - When target is reached
- **Notable events** - Interventions, policy changes
- **Unexpected changes** - Spikes or drops
- **Comparisons** - Differences from baseline

**Best Practices:**

1. **Don't over-annotate** - 2-4 annotations max
2. **Use arrows** for clarity
3. **Color-code** meaningfully (red = bad, green = good)
4. **Keep text brief** - Key numbers + interpretation
5. **Position carefully** - Avoid overlapping
:::
:::

## Small Multiples - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Compare many series side-by-side

**Goal:** 🔵 Evolution + 🟢 Ranking

```{r}
#| echo: true
#| eval: false

# Create small multiples by region
region_trends <- time_data %>%
  filter(!is.na(Region), !is.na(SysBP)) %>%
  group_by(Region, Month) %>%
  summarise(
    mean_bp = mean(SysBP, na.rm = TRUE),
    .groups = "drop"
  )

# Small multiples plot
ggplot(region_trends, aes(x = Month, y = mean_bp)) +
  geom_line(color = "#3C5488", linewidth = 1.2) +
  geom_point(size = 3, color = "#3C5488") +
  geom_area(fill = "#3C5488", alpha = 0.2) +
  facet_wrap(~ Region, ncol = 2) +  # Split by region
  labs(title = "Blood Pressure Trends Across Regions",
       subtitle = "Small multiples enable direct comparison",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom() +
  theme(
    strip.background = element_rect(fill = "gray90", color = "black"),
    strip.text = element_text(face = "bold", size = 11),
    panel.spacing = unit(1.5, "lines")  # Space between panels
  )
```
:::

## Small Multiples - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| message: false

region_trends <- time_data %>%
  filter(!is.na(Region), !is.na(SysBP)) %>%
  group_by(Region, Month) %>%
  summarise(
    mean_bp = mean(SysBP, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(region_trends, aes(x = Month, y = mean_bp)) +
  geom_line(color = "#3C5488", linewidth = 1.2) +
  geom_point(size = 3, color = "#3C5488") +
  geom_area(fill = "#3C5488", alpha = 0.2) +
  facet_wrap(~ Region, ncol = 2) +
  labs(title = "Blood Pressure Trends Across Regions",
       subtitle = "Small multiples enable direct comparison",
       x = "Time (months)",
       y = "Mean Systolic BP (mmHg)") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom() +
  theme(
    strip.background = element_rect(fill = "gray90", color = "black"),
    strip.text = element_text(face = "bold", size = 11),
    panel.spacing = unit(1.5, "lines")
  )
```

## Interpreting Small Multiples

::: {style="font-size: 0.9em;"}
**Power of Small Multiples:**

::: {.incremental}
1. **Same scale** enables direct comparison
   - Each region shows similar range

2. **Pattern differences** are immediately visible
   - West shows gradual increase
   - North shows decrease
   - East and South relatively stable

3. **Easy to spot outliers** or anomalies
   - West's upward trend is unusual

4. **Reduces cognitive load** vs. overlapping lines
   - Each series has its own space
   - No color confusion

**Insight:** Regional trends vary - need to investigate West region
:::
:::

## Percent Change Plot - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show relative changes from baseline

**Goal:** 🔵 Evolution + comparison

```{r}
#| echo: true
#| eval: false

# Calculate percent change from baseline
pct_change_data <- time_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP)) %>%
  group_by(TreatmentGroup, Month) %>%
  summarise(mean_bp = mean(SysBP, na.rm = TRUE), .groups = "drop") %>%
  group_by(TreatmentGroup) %>%
  mutate(
    baseline = first(mean_bp),           # Get baseline value
    pct_change = 100 * (mean_bp - baseline) / baseline  # % change
  )

# Create percent change plot
ggplot(pct_change_data, aes(x = Month, y = pct_change, 
                           color = TreatmentGroup, group = TreatmentGroup)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  labs(title = "Percent Change in Blood Pressure from Baseline",
       subtitle = "Negative values indicate improvement",
       x = "Time (months)",
       y = "Change from Baseline (%)",
       color = "Treatment") +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom()
```
:::

## Percent Change Plot - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false

pct_change_data <- time_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP)) %>%
  group_by(TreatmentGroup, Month) %>%
  summarise(mean_bp = mean(SysBP, na.rm = TRUE), .groups = "drop") %>%
  group_by(TreatmentGroup) %>%
  mutate(
    baseline = first(mean_bp),
    pct_change = 100 * (mean_bp - baseline) / baseline
  )

ggplot(pct_change_data, aes(x = Month, y = pct_change, 
                           color = TreatmentGroup, group = TreatmentGroup)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3.5) +
  labs(title = "Percent Change in Blood Pressure from Baseline",
       subtitle = "Negative values indicate improvement",
       x = "Time (months)",
       y = "Change from Baseline (%)",
       color = "Treatment") +
  scale_color_brewer(palette = "Set1") +
  scale_x_continuous(breaks = c(0, 6, 12)) +
  theme_custom()
```

## Interpreting Percent Change

::: {style="font-size: 0.9em;"}
**Why Use Percent Change?**

::: {.incremental}
1. **Normalizes different baselines**
   - Groups with different starting points comparable

2. **Shows relative effect size**
   - Treatment A: ~15% reduction
   - Treatment B: ~12% reduction
   - Control: ~0% change

3. **Makes communication easier**
   - "15% reduction" more intuitive than "20 mmHg decrease"

4. **Useful for different scales**
   - Can compare BP, cholesterol, weight, etc. on same plot

**Limitation:** Can mislead if baseline values are very small
:::
:::

## Slope Graph - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Highlight individual changes between two time points

**Goal:** 🔵 Evolution + individual patterns

```{r}
#| echo: true
#| eval: false

# Prepare data: baseline vs 12-month by treatment
slope_data <- health_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP_0m), !is.na(SysBP_6m)) %>%
  select(PatientID, TreatmentGroup, SysBP_0m, SysBP_6m) %>%
  group_by(TreatmentGroup) %>%
  slice_sample(n = 15) %>%  # Sample 15 patients per group
  ungroup() %>%
  pivot_longer(cols = c(SysBP_0m, SysBP_6m),
               names_to = "Time",
               values_to = "SysBP") %>%
  mutate(Time = ifelse(Time == "SysBP_0m", "Baseline", "6 Months"))

# Create slope graph
ggplot(slope_data, aes(x = Time, y = SysBP, group = PatientID,
                      color = TreatmentGroup)) +
  geom_line(alpha = 0.6, linewidth = 0.8) +
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~ TreatmentGroup, ncol = 3) +
  labs(title = "Individual Patient Blood Pressure Changes",
       subtitle = "Each line represents one patient (n=15 per group)",
       x = "",
       y = "Systolic BP (mmHg)") +
  scale_color_brewer(palette = "Dark2") +
  theme_custom() +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "gray90"),
        strip.text = element_text(face = "bold"))
```
:::

## Slope Graph - Output

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 6
#| message: false

set.seed(123)
slope_data <- health_data %>%
  filter(!is.na(TreatmentGroup), !is.na(SysBP_0m), !is.na(SysBP_6m)) %>%
  select(PatientID, TreatmentGroup, SysBP_0m, SysBP_6m) %>%
  group_by(TreatmentGroup) %>%
  slice_sample(n = 15) %>%
  ungroup() %>%
  pivot_longer(cols = c(SysBP_0m, SysBP_6m),
               names_to = "Time",
               values_to = "SysBP") %>%
  mutate(Time = ifelse(Time == "SysBP_0m", "Baseline", "6 Months"))

ggplot(slope_data, aes(x = Time, y = SysBP, group = PatientID,
                      color = TreatmentGroup)) +
  geom_line(alpha = 0.6, linewidth = 0.8) +
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~ TreatmentGroup, ncol = 3) +
  labs(title = "Individual Patient Blood Pressure Changes",
       subtitle = "Each line represents one patient (n=15 per group)",
       x = "",
       y = "Systolic BP (mmHg)") +
  scale_color_brewer(palette = "Dark2") +
  theme_custom() +
  theme(legend.position = "none",
        strip.background = element_rect(fill = "gray90"),
        strip.text = element_text(face = "bold"))
```

## Interpreting Slope Graphs

::: {style="font-size: 0.9em;"}
**What Slope Graphs Reveal:**

::: {.incremental}
1. **Treatment groups** show mostly downward slopes
   - Most patients improved

2. **Control group** shows mixed results
   - Some up, some down (natural variation)

3. **Individual variability** is visible
   - Not all patients respond equally
   - Some non-responders in treatment groups

4. **Response heterogeneity** matters
   - Need to identify why some don't respond

**Advantage:** Shows both group trend AND individual variation
:::
:::

## Chart Selection Guide

::: {style="font-size: 0.8em;"}
| Chart Type | Goal | Best For | Complexity |
|------------|------|----------|------------|
| **LOESS Smooth** | 🔵 Evolution | Noisy data, reveal trends | Medium |
| **Area + CI** | 🔵 + 🟡 | Show uncertainty over time | Medium |
| **Annotated** | 🔵 + context | Presentations, reports | Easy-Medium |
| **Small Multiples** | 🔵 + 🟢 | Compare many series | Easy |
| **Percent Change** | 🔵 + comparison | Different baselines | Easy |
| **Slope Graph** | 🔵 + individuals | Before/after comparisons | Easy |

**Pro Tip:** Combine techniques - e.g., small multiples with smoothing
:::

## Common Mistakes

::: {.incremental style="font-size: 0.9em;"}
1. **Over-smoothing** - Losing important details
   - Solution: Try different span values

2. **Too many lines** on one plot
   - Solution: Use small multiples or filter

3. **Ignoring uncertainty**
   - Solution: Always show confidence intervals

4. **Poor annotation** placement
   - Solution: Use arrow() and adjust positions

5. **Wrong baseline** for percent change
   - Solution: Clearly document reference point

6. **Missing time labels**
   - Solution: Use clear date formats or labels
:::

## Practice Exercise

**Your Turn:**

1. Create a LOESS smoothed plot of mean `Cholesterol` over time by `Gender`
2. Add confidence bands to the plot
3. Create a slope graph showing individual `BMI` changes (0m → 12m) by `ExerciseFreq`
4. Build small multiples showing `BloodSugar` trends across all 4 Regions

**Bonus Challenge:**

Create an annotated time series highlighting when Treatment A achieved target BP (<120 mmHg).

## Summary

::: {style="font-size: 0.9em;"}
**What You've Learned:**

::: {.incremental}
1. **LOESS smoothing** reveals underlying trends
2. **Confidence bands** show uncertainty
3. **Annotations** add context and insights
4. **Small multiples** enable easy comparison
5. **Percent change** normalizes baselines
6. **Slope graphs** show individual trajectories

**Key Principle:**

> Time series plots should tell a story - show the trend, highlight changes, and provide context
:::
:::

## Next Steps {.center}

**You've mastered Time Series - Intermediate!**

**Next:** Relational/Flow - Intermediate

Learn to visualize relationships with:

- Network diagrams
- Sankey diagrams
- Chord diagrams
- Alluvial plots
