---
title: "Relational/Flow - INTERMEDIATE"
subtitle: "Visualizing Connections and Flows"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Relational & Flow Data {.center}

## What We'll Learn - Intermediate

::: {.columns}
::: {.column width="50%"}
**Visualizing Relationships:**

- Network diagrams (nodes & edges)
- Sankey diagrams (flow)
- Chord diagrams (circular relationships)
- Alluvial plots (categorical flows)
- Correlation networks
:::

::: {.column width="50%"}
**Analytical Goals:**

- 🔴 **Flow** - Movement between states
- ⚫ **Correlation** - Strength of relationships
- 🟢 **Ranking** - Most connected nodes
- 🟡 **Distribution** - Network properties
:::
:::

## Setup Code

```{r}
#| echo: true
#| message: false

# Load required packages
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(igraph)      # For network analysis
library(ggraph)      # For network visualization
library(networkD3)   # For Sankey diagrams
library(circlize)    # For chord diagrams

# Load datasets
health_data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")
network_data <- read_excel("../../data/Healthcare_Network_Dataset.xlsx", sheet = 2)

# Create custom theme
theme_custom <- function() {
  theme_void() +  # Clean background for networks
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "right"
  )
}
```

## Network Diagram Basics - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show relationships between entities (nodes)

**Goal:** ⚫ Correlation + 🟢 Ranking

```{r}
#| echo: true
#| eval: false

# Prepare network data (patient referrals between providers)
edges <- network_data %>%
  select(from = FromProvider, to = ToProvider, 
           weight = NumReferrals) %>%
  filter(!is.na(from), !is.na(to), weight > 0)

# Create network graph object
network_graph <- graph_from_data_frame(edges, directed = TRUE)

# Calculate node importance (degree centrality)
V(network_graph)$degree <- degree(network_graph)
V(network_graph)$betweenness <- betweenness(network_graph)

# Plot network
ggraph(network_graph, layout = "fr") +  # Fruchterman-Reingold layout
  geom_edge_link(aes(width = weight, alpha = weight),
                arrow = arrow(length = unit(3, "mm")),
                end_cap = circle(3, "mm"),
                color = "gray60") +
  geom_node_point(aes(size = degree, color = degree)) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_edge_width(range = c(0.5, 3)) +
  scale_edge_alpha(range = c(0.3, 0.8)) +
  scale_size(range = c(3, 10)) +
  scale_color_viridis_c(option = "plasma") +
  labs(title = "Healthcare Provider Referral Network",
       subtitle = "Node size = number of connections | Arrow width = referral volume",
       size = "Connections",
       color = "Connections") +
  theme_custom()
```
:::

## Network Diagram - Output

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| warning: false
#| message: false

edges <- network_data %>%
  select(from = FromProvider, to = ToProvider, 
         weight = NumReferrals) %>%
  filter(!is.na(from), !is.na(to), weight > 0)

network_graph <- graph_from_data_frame(edges, directed = TRUE)
V(network_graph)$degree <- igraph::degree(network_graph)
V(network_graph)$betweenness <- betweenness(network_graph)

set.seed(123)
ggraph(network_graph, layout = "fr") +
  geom_edge_link(aes(width = weight, alpha = weight),
                arrow = arrow(length = unit(3, "mm")),
                end_cap = circle(3, "mm"),
                color = "gray60") +
  geom_node_point(aes(size = degree, color = degree)) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3) +
  scale_edge_width(range = c(0.5, 3)) +
  scale_edge_alpha(range = c(0.3, 0.8)) +
  scale_size(range = c(3, 10)) +
  scale_color_viridis_c(option = "plasma") +
  labs(title = "Healthcare Provider Referral Network",
       subtitle = "Node size = number of connections | Arrow width = referral volume",
       size = "Connections",
       color = "Connections") +
  theme_custom()
```

## Interpreting Network Diagrams

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**Network Metrics:**

- **Node size** - Degree centrality (connections)
- **Node color** - Importance measure
- **Edge width** - Connection strength
- **Edge direction** - Flow direction

**Key Nodes:**

- **Hubs** - Many connections (large nodes)
- **Bridges** - Connect clusters
- **Periphery** - Few connections
:::

::: {.column width="50%"}
**Our Network:**

1. **Central hubs identified**
   - Key providers with most referrals

2. **Clusters visible**
   - Groups of closely connected providers
   - Possible subspecialty groups

3. **Isolated providers** exist
   - May need better integration

4. **Referral patterns** show hierarchy
   - Primary care → specialists
   - Direction matters
:::
:::

## Sankey Diagram - Code

::: {style="font-size: 0.75em;"}
**Purpose:** Show flow quantities between stages

**Goal:** 🔴 Flow + 🔴 Part of Whole

```{r}
#| echo: true
#| eval: false

# Prepare flow data: Patient journey through treatment
# Stage 1: Initial Assessment → Stage 2: Treatment Selection → Stage 3: Outcome

# Create flow data
flow_data <- health_data %>%
  filter(!is.na(TreatmentGroup), !is.na(Adherence_Status)) %>%
  count(TreatmentGroup, Adherence_Status, name = "count") %>%
  # Rename for Sankey format
  rename(source = TreatmentGroup, target = Adherence_Status, value = count)

# Prepare nodes and links for networkD3
nodes <- data.frame(
  name = unique(c(as.character(flow_data$source), 
                 as.character(flow_data$target)))
)

# Create links with node indices
flow_data$IDsource <- match(flow_data$source, nodes$name) - 1  # 0-indexed
flow_data$IDtarget <- match(flow_data$target, nodes$name) - 1

# Create Sankey diagram
sankeyNetwork(
  Links = flow_data,
  Nodes = nodes,
  Source = "IDsource",
  Target = "IDtarget",
  Value = "value",
  NodeID = "name",
  units = "patients",
  fontSize = 14,
  nodeWidth = 30,
  sinksRight = FALSE,  # Align target nodes to left
  colourScale = JS('d3.scaleOrdinal().domain(["Treatment A", "Treatment B", "Control", "High", "Low"]).range(["#E41A1C", "#377EB8", "#4DAF4A", "#FF7F00", "#984EA3"])')
)
```

**Note:** This creates an interactive HTML widget
:::

## Sankey Diagram - Interpretation

::: {style="font-size: 0.9em;"}
**Reading Sankey Diagrams:**

::: {.incremental}
1. **Width of flow** = Quantity
   - Wider flows = more patients

2. **Color** helps track paths
   - Each treatment group has its own color

3. **Convergence/Divergence** shows patterns
   - Where do patients end up?
   - Which paths are most common?

**Our Findings:**

- **Treatment A & B** show higher adherence
  - Larger flows to "High" adherence
- **Control group** has lower adherence
  - More patients in "Low" adherence
- **Visual proportions** immediately show treatment effectiveness
:::
:::

## Alluvial Plot - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show changes in categorical composition over time

**Goal:** 🔴 Flow + 🟡 Distribution

```{r}
#| echo: true
#| eval: false

library(ggalluvial)

# Prepare data: Smoking status changes over study
alluvial_data <- health_data %>%
  select(PatientID, SmokingStatus, Gender, TreatmentGroup) %>%
  filter(!is.na(SmokingStatus), !is.na(Gender), !is.na(TreatmentGroup)) %>%
  count(TreatmentGroup, SmokingStatus, Gender)

# Create alluvial plot
ggplot(alluvial_data,
       aes(y = n, axis1 = TreatmentGroup, axis2 = SmokingStatus, 
           axis3 = Gender)) +
  geom_alluvium(aes(fill = SmokingStatus), width = 1/12, alpha = 0.7) +
  geom_stratum(width = 1/12, fill = "gray90", color = "black") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
  scale_x_discrete(limits = c("Treatment", "Smoking", "Gender"),
                   expand = c(0.05, 0.05)) +
  scale_fill_manual(values = c("Never" = "#00A087", 
                              "Former" = "#3C5488",
                              "Current" = "#DC0000")) +
  labs(title = "Patient Flow: Treatment → Smoking Status → Gender",
       subtitle = "Stream width shows number of patients in each pathway",
       y = "Number of Patients",
       fill = "Smoking Status") +
  theme_minimal() +
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.grid = element_blank()
  )
```
:::

## Alluvial Plot - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| warning: false
#| message: false

library(ggalluvial)

alluvial_data <- health_data %>%
  select(PatientID, SmokingStatus, Gender, TreatmentGroup) %>%
  filter(!is.na(SmokingStatus), !is.na(Gender), !is.na(TreatmentGroup)) %>%
  count(TreatmentGroup, SmokingStatus, Gender)

ggplot(alluvial_data,
       aes(y = n, axis1 = TreatmentGroup, axis2 = SmokingStatus, 
           axis3 = Gender)) +
  geom_alluvium(aes(fill = SmokingStatus), width = 1/12, alpha = 0.7) +
  geom_stratum(width = 1/12, fill = "gray90", color = "black") +
  geom_label(stat = "stratum", aes(label = after_stat(stratum)), size = 3) +
  scale_x_discrete(limits = c("Treatment", "Smoking", "Gender"),
                   expand = c(0.05, 0.05)) +
  scale_fill_manual(values = c("Never" = "#00A087", 
                              "Former" = "#3C5488",
                              "Current" = "#DC0000")) +
  labs(title = "Patient Flow: Treatment → Smoking Status → Gender",
       subtitle = "Stream width shows number of patients in each pathway",
       y = "Number of Patients",
       fill = "Smoking Status") +
  theme_minimal() +
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    axis.text.y = element_text(size = 10),
    panel.grid = element_blank()
  )
```

## Interpreting Alluvial Plots

::: {style="font-size: 0.9em;"}
**Key Features:**

::: {.incremental}
1. **Stream thickness** = Quantity flowing through that path
   - Follow a color to track one category

2. **Convergence** shows common endpoints
   - Many paths leading to same outcome

3. **Divergence** shows splits
   - One group splits into multiple outcomes

**Our Insights:**

- **Never smokers** largest group in all treatments
- **Gender distribution** fairly balanced across smoking statuses
- **Treatment allocation** appears balanced across smoking groups
- **Visual flow** shows patient composition clearly
:::
:::

## Chord Diagram - Code {.scrollable}

::: {style="font-size: 0.75em;"}
**Purpose:** Show relationships between categories in circular layout

**Goal:** ⚫ Correlation + 🔴 Flow

```{r}
#| echo: true
#| eval: false

# Prepare chord data: Comorbidity co-occurrence
# Simplified: Which regions refer to which treatment groups most

chord_data <- network_data %>%
  filter(!is.na(PatientRegion), !is.na(TreatmentAssigned)) %>%
  count(PatientRegion, TreatmentAssigned, name = "count") %>%
  pivot_wider(names_from = TreatmentAssigned, 
              values_from = count, 
              values_fill = 0)

# Convert to matrix
chord_matrix <- as.matrix(chord_data[, -1])
rownames(chord_matrix) <- chord_data$PatientRegion

# Create chord diagram
circos.clear()
circos.par(start.degree = 90, gap.degree = 4)

chordDiagram(
  chord_matrix,
  grid.col = c("North" = "#E41A1C", "South" = "#377EB8", 
               "East" = "#4DAF4A", "West" = "#984EA3",
               "Treatment A" = "#FF7F00", "Treatment B" = "#FFFF33", 
               "Control" = "#A65628"),
  transparency = 0.5,
  directional = 1,
  direction.type = c("arrows", "diffHeight"),
  diffHeight = -0.04,
  annotationTrack = "grid",
  annotationTrackHeight = c(0.05, 0.1),
  link.arr.type = "big.arrow",
  link.sort = TRUE,
  link.largest.ontop = TRUE
)

# Add labels
circos.trackPlotRegion(
  track.index = 1,
  bg.border = NA,
  panel.fun = function(x, y) {
    sector.index = get.cell.meta.data("sector.index")
    circos.text(CELL_META$xcenter, CELL_META$ylim[1], sector.index,
               facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5))
  }
)
```
:::

## Chord Diagram - Interpretation

::: {style="font-size: 0.9em;"}
**Reading Chord Diagrams:**

::: {.incremental}
1. **Arcs** represent categories
   - Arc width = total connections

2. **Ribbons** show connections
   - Ribbon width = flow strength
   - Color indicates source

3. **Circular layout** saves space
   - Shows all connections simultaneously

**Our Pattern:**

- **East region** sends most patients to Treatment A
- **North region** more balanced across treatments
- **West region** has more Control assignments
- **Treatment allocation** varies by region (interesting!)
:::

**When to use:** Many-to-many relationships, limited space
:::

## Correlation Network - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Visualize correlation patterns as network

**Goal:** ⚫ Correlation + clustering

```{r}
#| echo: true
#| eval: false

# Calculate correlations among numeric health metrics
cor_data <- health_data %>%
  select(Age, BMI_0m, SysBP_0m, DiaBP_0m, 
         Cholesterol_0m, BloodSugar_0m) %>%
  cor(use = "complete.obs")

# Convert to network (only strong correlations)
cor_network <- cor_data %>%
  as.data.frame() %>%
  mutate(from = rownames(.)) %>%
  pivot_longer(cols = -from, names_to = "to", values_to = "correlation") %>%
  filter(from != to, abs(correlation) > 0.3) %>%  # Threshold
  mutate(abs_cor = abs(correlation))

# Create graph
cor_graph <- graph_from_data_frame(cor_network, directed = FALSE)
E(cor_graph)$weight <- cor_network$abs_cor
E(cor_graph)$cor_sign <- ifelse(cor_network$correlation > 0, "Positive", "Negative")

# Plot
ggraph(cor_graph, layout = "fr") +
  geom_edge_link(aes(width = abs_cor, color = cor_sign), alpha = 0.7) +
  geom_node_point(size = 10, color = "#4DBBD5") +
  geom_node_text(aes(label = name), size = 3.5, color = "white", fontface = "bold") +
  scale_edge_width(range = c(1, 4)) +
  scale_edge_color_manual(values = c("Positive" = "#00A087", "Negative" = "#DC0000")) +
  labs(title = "Health Metrics Correlation Network",
       subtitle = "Edge width = correlation strength | Color = positive/negative",
       edge_color = "Correlation",
       edge_width = "Strength") +
  theme_custom()
```
:::

## Correlation Network - Output

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 7
#| warning: false

cor_data <- health_data %>%
  select(Age, BMI, SysBP_0m, DiaBP_0m, 
         Cholesterol, BloodSugar) %>%
  cor(use = "complete.obs")

cor_network <- cor_data %>%
  as.data.frame() %>%
  mutate(from = rownames(.)) %>%
  pivot_longer(cols = -from, names_to = "to", values_to = "correlation") %>%
  filter(from != to, abs(correlation) > 0.3) %>%
  mutate(abs_cor = abs(correlation))

cor_graph <- graph_from_data_frame(cor_network, directed = FALSE)
E(cor_graph)$weight <- cor_network$abs_cor
E(cor_graph)$cor_sign <- ifelse(cor_network$correlation > 0, "Positive", "Negative")

set.seed(456)
ggraph(cor_graph, layout = "fr") +
  geom_edge_link(aes(width = abs_cor, color = cor_sign), alpha = 0.7) +
  geom_node_point(size = 10, color = "#4DBBD5") +
  geom_node_text(aes(label = name), size = 3.5, color = "white", fontface = "bold") +
  scale_edge_width(range = c(1, 4)) +
  scale_edge_color_manual(values = c("Positive" = "#00A087", "Negative" = "#DC0000")) +
  labs(title = "Health Metrics Correlation Network",
       subtitle = "Edge width = correlation strength | Color = positive/negative",
       edge_color = "Correlation",
       edge_width = "Strength") +
  theme_custom()
```

## Interpreting Correlation Networks

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**Advantages:**

- **Clusters** emerge naturally
  - Related variables group together

- **Strength** shown by thickness
  - Thicker = stronger correlation

- **Direction** shown by color
  - Green = positive
  - Red = negative

- **Hub variables** visible
  - Most connected = most important
:::

::: {.column width="50%"}
**Our Network:**

1. **SysBP & DiaBP** tightly linked
   - Expected (measured together)

2. **BMI cluster** with BP measures
   - Weight affects blood pressure

3. **Blood sugar** moderately connected
   - Related to metabolic health

4. **Age** shows weaker connections
   - Less directly correlated with others
:::
:::

## Chart Selection Guide

::: {style="font-size: 0.85em;"}
| Chart Type | Goal | Best For | Complexity |
|------------|------|----------|------------|
| **Network Diagram** | ⚫ Correlation + 🟢 | Show connections, identify hubs | High |
| **Sankey Diagram** | 🔴 Flow | Show quantity flows between stages | Medium |
| **Alluvial Plot** | 🔴 Flow + 🟡 | Categorical changes over stages | Medium |
| **Chord Diagram** | ⚫ + 🔴 Flow | Many-to-many relationships | High |
| **Correlation Network** | ⚫ Correlation | Show variable relationships | Medium-High |

**Pro Tip:** Choose based on data structure and story you want to tell
:::

## Common Mistakes

::: {.incremental style="font-size: 0.9em;"}
1. **Too many nodes/edges**
   - Solution: Filter to top connections, increase threshold

2. **Poor layout** choice
   - Solution: Try different layouts (fr, kk, circle, tree)

3. **Unreadable labels**
   - Solution: Use `repel = TRUE`, adjust size, or interactive plots

4. **Ignoring direction** in networks
   - Solution: Use arrows for directed graphs

5. **Color overload** in complex diagrams
   - Solution: Limit to 2-3 colors, use grayscale for edges

6. **Not showing flow magnitude** clearly
   - Solution: Vary width, transparency, or both
:::

## Interactive Networks

::: {style="font-size: 0.9em;"}
**Benefits of Interactive Plots:**

```{r}
#| echo: true
#| eval: false

library(visNetwork)

# Prepare data for interactive network
nodes_df <- data.frame(
  id = V(network_graph)$name,
  label = V(network_graph)$name,
  value = V(network_graph)$degree,  # Size by degree
  title = paste0(V(network_graph)$name, "<br>",
                "Connections: ", V(network_graph)$degree)  # Tooltip
)

edges_df <- as_data_frame(network_graph) %>%
  rename(from = from, to = to) %>%
  mutate(value = weight,  # Width by weight
         title = paste0("Referrals: ", weight))  # Tooltip

# Create interactive network
visNetwork(nodes_df, edges_df) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visPhysics(stabilization = TRUE)
```

**Features:** Hover tooltips, zoom, drag nodes, highlight neighbors
:::

## Practice Exercise

**Your Turn:**

1. Create a network diagram showing correlations between `Cholesterol`, `BMI`, `BloodSugar`, and `Age`
2. Build a Sankey diagram showing flow from `Region` → `TreatmentGroup` → `Adherence_Status`
3. Create an alluvial plot for `Gender` → `ExerciseFreq` → `SmokingStatus`

**Bonus Challenge:**

Create a chord diagram showing referral patterns between different provider specialties (if you have specialty data).

## Summary

::: {style="font-size: 0.9em;"}
**What You've Learned:**

::: {.incremental}
1. **Network diagrams** reveal connection patterns and hubs
2. **Sankey diagrams** show quantity flows between stages
3. **Alluvial plots** visualize categorical changes
4. **Chord diagrams** display many-to-many relationships
5. **Correlation networks** identify variable clusters

**Key Principle:**

> Relational visualizations reveal hidden patterns in connections, flows, and associations
:::
:::

## Next Steps {.center}

**You've mastered Relational/Flow - Intermediate!**

**Next:** Geographic - Intermediate

Learn spatial visualization with:

- Choropleth maps
- Point maps with clustering
- Heat maps
- Faceted maps
