---
title: "Geographic - INTERMEDIATE"
subtitle: "Spatial Data Visualization"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Geographic Visualization {.center}

## What We'll Learn - Intermediate

::: {.columns}
::: {.column width="50%"}
**Spatial Data Visualization:**

- Choropleth maps (filled regions)
- Point/bubble maps
- Heat maps (density)
- Faceted maps (small multiples)
- Interactive maps
:::

::: {.column width="50%"}
**Analytical Goals:**

- 🟣 **Geographic** - Where things happen
- 🟡 **Distribution** - Spatial patterns
- 🟢 **Ranking** - Regional comparisons
- ⚫ **Correlation** - Geographic relationships
:::
:::

## Setup Code

```{r}
#| echo: true
#| message: false

# Load required packages
library(readxl)
library(tidyverse)
library(ggplot2)
library(sf)          # For spatial data
library(maps)        # For map data
library(viridis)     # For color scales

# Load dataset
health_data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Aggregate data by region
region_summary <- health_data %>%
  filter(!is.na(Region)) %>%
  group_by(Region) %>%
  summarise(
    mean_bp = mean(SysBP_0m, na.rm = TRUE),
    mean_cholesterol = mean(Cholesterol, na.rm = TRUE),
    mean_bmi = mean(BMI, na.rm = TRUE),
    n_patients = n(),
    hypertension_rate = mean(SysBP_0m >= 140, na.rm = TRUE) * 100
  )

# Create theme
theme_map <- function() {
  theme_void() +
  theme(
    text = element_text(family = "serif", size = 12),
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 10),
    panel.background = element_rect(fill = "aliceblue")
  )
}
```

## Choropleth Map - Code

::: {style="font-size: 0.75em;"}
**Purpose:** Show values across geographic regions using color

**Goal:** 🟣 Geographic + 🟢 Ranking

```{r}
#| echo: true
#| eval: false

# Get US state map data
us_states <- map_data("state")

# Create region to state mapping (simplified for demonstration)
region_states <- data.frame(
  region_name = c("North", "South", "East", "West"),
  states = c("new york,pennsylvania,ohio,michigan",
            "texas,florida,georgia,alabama",
            "massachusetts,connecticut,rhode island,maine",
            "california,oregon,washington,nevada")
)

# Expand and join
state_regions <- region_states %>%
  separate_rows(states, sep = ",") %>%
  rename(region_lower = states)

# Join with map data
map_data_full <- us_states %>%
  left_join(state_regions, by = c("region" = "region_lower")) %>%
  left_join(region_summary, by = c("region_name" = "Region"))

# Create choropleth map
ggplot(map_data_full, aes(x = long, y = lat, group = group, fill = mean_bp)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_map("albers", lat0 = 39, lat1 = 45) +  # Albers projection
  scale_fill_viridis_c(
    option = "plasma",
    name = "Mean Systolic BP\n(mmHg)",
    na.value = "gray90",
    guide = guide_colorbar(
      barwidth = 1.5,
      barheight = 15,
      title.position = "top"
    )
  ) +
  labs(title = "Average Blood Pressure by Region",
       subtitle = "Darker colors indicate higher blood pressure",
       caption = "Data: PublicHealth BP Dataset") +
  theme_map()
```
:::

## Choropleth Map - Output

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| warning: false
#| message: false

us_states <- map_data("state")

region_states <- data.frame(
  region_name = c("North", "South", "East", "West"),
  states = c("new york,pennsylvania,ohio,michigan",
            "texas,florida,georgia,alabama",
            "massachusetts,connecticut,rhode island,maine",
            "california,oregon,washington,nevada")
)

state_regions <- region_states %>%
  separate_rows(states, sep = ",") %>%
  rename(region_lower = states)

map_data_full <- us_states %>%
  left_join(state_regions, by = c("region" = "region_lower")) %>%
  left_join(region_summary, by = c("region_name" = "Region"))

ggplot(map_data_full, aes(x = long, y = lat, group = group, fill = mean_bp)) +
  geom_polygon(color = "white", size = 0.3) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_fill_viridis_c(
    option = "plasma",
    name = "Mean Systolic BP\n(mmHg)",
    na.value = "gray90",
    guide = guide_colorbar(
      barwidth = 1.5,
      barheight = 15,
      title.position = "top"
    )
  ) +
  labs(title = "Average Blood Pressure by Region",
       subtitle = "Darker colors indicate higher blood pressure",
       caption = "Data: PublicHealth BP Dataset") +
  theme_map()
```

## Interpreting Choropleth Maps

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**Key Features:**

- **Color intensity** = Value magnitude
- **Patterns** reveal regional clustering
- **White borders** separate regions
- **Gray** indicates missing data

**Best Practices:**

- Use sequential color scales for continuous data
- Use diverging scales for data with meaningful midpoint
- Limit to 5-7 color bins for categorical
- Always include legend with units
:::

::: {.column width="50%"}
**Our Map Shows:**

1. **Geographic variation** in BP
   - Clear regional differences

2. **Hot spots** identified
   - East region shows highest BP

3. **Cool spots** visible
   - West region shows lowest BP

4. **Spatial patterns** suggest
   - Possible lifestyle/environmental factors
   - Need for targeted interventions
:::
:::

## Point/Bubble Map - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show individual locations with value-sized markers

**Goal:** 🟣 Geographic + 🟡 Distribution

```{r}
#| echo: true
#| eval: false

# Create sample patient locations (simulated for demonstration)
set.seed(123)
patient_locations <- health_data %>%
  filter(!is.na(Region), !is.na(SysBP_0m)) %>%
  mutate(
    # Simulate coordinates within regions
    lat = case_when(
      Region == "North" ~ runif(n(), 40, 45),
      Region == "South" ~ runif(n(), 30, 35),
      Region == "East" ~ runif(n(), 39, 43),
      Region == "West" ~ runif(n(), 35, 42)
    ),
    lon = case_when(
      Region == "North" ~ runif(n(), -85, -75),
      Region == "South" ~ runif(n(), -100, -80),
      Region == "East" ~ runif(n(), -75, -70),
      Region == "West" ~ runif(n(), -125, -115)
    ),
    hypertensive = ifelse(SysBP_0m >= 140, "Yes", "No")
  ) %>%
  sample_n(200)  # Sample for clarity

# Create point map
ggplot() +
  geom_polygon(data = us_states, 
               aes(x = long, y = lat, group = group),
               fill = "white", color = "gray70") +
  geom_point(data = patient_locations,
            aes(x = lon, y = lat, color = hypertensive, size = SysBP_0m),
            alpha = 0.6) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_color_manual(values = c("No" = "#00A087", "Yes" = "#DC0000"),
                    name = "Hypertensive") +
  scale_size_continuous(range = c(1, 6), name = "Systolic BP\n(mmHg)") +
  labs(title = "Patient Locations by Blood Pressure Status",
       subtitle = "Point size = BP value | Color = Hypertension status (≥140 mmHg)",
       caption = "Sample of 200 patients") +
  theme_map()
```
:::

## Point/Bubble Map - Output

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| warning: false
#| message: false

set.seed(123)
patient_locations <- health_data %>%
  filter(!is.na(Region), !is.na(SysBP_0m)) %>%
  mutate(
    lat = case_when(
      Region == "North" ~ runif(n(), 40, 45),
      Region == "South" ~ runif(n(), 30, 35),
      Region == "East" ~ runif(n(), 39, 43),
      Region == "West" ~ runif(n(), 35, 42)
    ),
    lon = case_when(
      Region == "North" ~ runif(n(), -85, -75),
      Region == "South" ~ runif(n(), -100, -80),
      Region == "East" ~ runif(n(), -75, -70),
      Region == "West" ~ runif(n(), -125, -115)
    ),
    hypertensive = ifelse(SysBP_0m >= 140, "Yes", "No")
  ) %>%
  sample_n(200)

ggplot() +
  geom_polygon(data = us_states, 
               aes(x = long, y = lat, group = group),
               fill = "white", color = "gray70") +
  geom_point(data = patient_locations,
            aes(x = lon, y = lat, color = hypertensive, size = SysBP_0m),
            alpha = 0.6) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_color_manual(values = c("No" = "#00A087", "Yes" = "#DC0000"),
                    name = "Hypertensive") +
  scale_size_continuous(range = c(1, 6), name = "Systolic BP\n(mmHg)") +
  labs(title = "Patient Locations by Blood Pressure Status",
       subtitle = "Point size = BP value | Color = Hypertension status (≥140 mmHg)",
       caption = "Sample of 200 patients") +
  theme_map()
```

## Interpreting Point/Bubble Maps

::: {style="font-size: 0.9em;"}
**What Point Maps Show:**

::: {.incremental}
1. **Exact locations** of events/observations
   - Each point = one patient

2. **Clustering patterns**
   - Where do high-BP patients concentrate?
   - Red dots (hypertensive) cluster in certain areas

3. **Spatial distribution**
   - Uniform vs. clustered
   - Our data shows regional clustering

4. **Multiple attributes** simultaneously
   - Size = BP value
   - Color = hypertension status
   - Position = location

**Advantage:** Shows individual variation AND spatial pattern
:::
:::

## Heat Map (Density) - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Show density/concentration of observations

**Goal:** 🟣 Geographic + 🟡 Distribution

```{r}
#| echo: true
#| eval: false

# Create density heat map
ggplot() +
  # Base map
  geom_polygon(data = us_states, 
               aes(x = long, y = lat, group = group),
               fill = "gray95", color = "white") +
  # Density contours
  stat_density_2d(data = patient_locations,
                 aes(x = lon, y = lat, fill = after_stat(level)),
                 geom = "polygon",
                 alpha = 0.5,
                 bins = 10) +
  # Overlay points
  geom_point(data = patient_locations,
            aes(x = lon, y = lat),
            size = 0.5, alpha = 0.3, color = "black") +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_fill_viridis_c(option = "inferno",
                      name = "Patient\nDensity") +
  labs(title = "Patient Density Heat Map",
       subtitle = "Darker areas show higher patient concentration",
       caption = "Contours show density levels") +
  theme_map()
```
:::

## Heat Map (Density) - Output

```{r}
#| echo: false
#| fig-width: 11
#| fig-height: 7
#| warning: false
#| message: false

ggplot() +
  geom_polygon(data = us_states, 
               aes(x = long, y = lat, group = group),
               fill = "gray95", color = "white") +
  stat_density_2d(data = patient_locations,
                 aes(x = lon, y = lat, fill = after_stat(level)),
                 geom = "polygon",
                 alpha = 0.5,
                 bins = 10) +
  geom_point(data = patient_locations,
            aes(x = lon, y = lat),
            size = 0.5, alpha = 0.3, color = "black") +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_fill_viridis_c(option = "inferno",
                      name = "Patient\nDensity") +
  labs(title = "Patient Density Heat Map",
       subtitle = "Darker areas show higher patient concentration",
       caption = "Contours show density levels") +
  theme_map()
```

## Interpreting Heat Maps

::: {.columns style="font-size: 0.9em;"}
::: {.column width="50%"}
**Heat Map Features:**

- **Color intensity** = Density
- **Contour lines** = Density levels
- **Hot spots** = High concentration
- **Cool areas** = Low concentration

**When to Use:**

- Many overlapping points
- Want to show concentration
- Exact locations less important than density
- Identifying high-activity areas
:::

::: {.column width="50%"}
**Our Heat Map:**

1. **Clear hot spots** in East
   - Highest patient density

2. **Moderate density** in North & South
   - Evenly distributed

3. **Sparse areas** in West
   - Lower patient numbers

4. **Urban/rural patterns** visible
   - Concentration suggests urban centers

**Insight:** Resources should focus on high-density areas
:::
:::

## Faceted Maps - Code

::: {style="font-size: 0.8em;"}
**Purpose:** Compare spatial patterns across categories

**Goal:** 🟣 Geographic + 🟢 Comparison

```{r}
#| echo: true
#| eval: false

# Create faceted map
ggplot() +
  geom_polygon(data = us_states, 
               aes(x = long, y = lat, group = group),
               fill = "white", color = "gray70") +
  geom_point(data = patient_locations,
            aes(x = lon, y = lat, color = hypertensive),
            size = 2, alpha = 0.6) +
  facet_wrap(~ TreatmentGroup, ncol = 2) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_color_manual(values = c("No" = "#00A087", "Yes" = "#DC0000"),
                    name = "Hypertensive") +
  labs(title = "Patient Distribution by Treatment Group",
       subtitle = "Geographic distribution across treatment assignments",
       caption = "Each panel shows one treatment group") +
  theme_map() +
  theme(
    strip.background = element_rect(fill = "gray90", color = "black"),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  )
```
:::

## Faceted Maps - Output

```{r}
#| echo: false
#| fig-width: 12
#| fig-height: 8
#| warning: false
#| message: false


ggplot() +
  geom_polygon(data = us_states, 
               aes(x = long, y = lat, group = group),
               fill = "white", color = "gray70") +
  geom_point(data = patient_locations,
            aes(x = lon, y = lat, color = hypertensive),
            size = 2, alpha = 0.6) +
  facet_wrap(~ TreatmentGroup, ncol = 2) +
  coord_map("albers", lat0 = 39, lat1 = 45) +
  scale_color_manual(values = c("No" = "#00A087", "Yes" = "#DC0000"),
                    name = "Hypertensive") +
  labs(title = "Patient Distribution by Treatment Group",
       subtitle = "Geographic distribution across treatment assignments",
       caption = "Each panel shows one treatment group") +
  theme_map() +
  theme(
    strip.background = element_rect(fill = "gray90", color = "black"),
    strip.text = element_text(face = "bold", size = 10),
    legend.position = "bottom"
  )
```

## Interpreting Faceted Maps

::: {style="font-size: 0.9em;"}
**Power of Faceted Maps:**

::: {.incremental}
1. **Direct comparison** across categories
   - Same scale enables comparison

2. **Pattern differences** immediately visible
   - Treatment allocation varies by region

3. **Consistency checks**
   - Is treatment assignment geographically balanced?

4. **Multiple variables** displayed
   - Location + Treatment + Hypertension status

**Our Finding:**

- Treatment groups fairly balanced geographically
- Hypertension prevalence similar across groups
- Good randomization evident
:::
:::

## Interactive Maps (Conceptual)

::: {style="font-size: 0.9em;"}
**Using Leaflet for Interactive Maps:**

```{r}
#| echo: true
#| eval: false

library(leaflet)

# Create interactive map
leaflet(patient_locations) %>%
  addTiles() %>%  # Add OpenStreetMap tiles
  addCircleMarkers(
    ~lon, ~lat,
    radius = ~SysBP_0m / 20,  # Size by BP
    color = ~ifelse(hypertensive == "Yes", "red", "green"),
    fillOpacity = 0.6,
    popup = ~paste0(
      "<b>Patient ID:</b> ", PatientID, "<br>",
      "<b>Systolic BP:</b> ", SysBP_0m, " mmHg<br>",
      "<b>Region:</b> ", Region, "<br>",
      "<b>Hypertensive:</b> ", hypertensive
    )
  ) %>%
  addLegend(
    position = "bottomright",
    colors = c("green", "red"),
    labels = c("Normal BP", "Hypertensive"),
    title = "BP Status"
  )
```

**Features:** Zoom, pan, click for details, layer toggle
:::

## Chart Selection Guide

::: {style="font-size: 0.85em;"}
| Chart Type | Goal | Best For | Data Type |
|------------|------|----------|-----------|
| **Choropleth** | 🟣 + 🟢 | Regional comparisons | Aggregated by region |
| **Point/Bubble** | 🟣 + 🟡 | Individual locations | Point data |
| **Heat Map** | 🟣 + 🟡 | Density patterns | Many overlapping points |
| **Faceted Maps** | 🟣 + 🟢 | Compare across categories | Point or regional data |
| **Interactive** | All | Exploration, presentations | Any spatial data |

**Pro Tip:** Consider your audience - static for reports, interactive for exploration
:::

## Common Mistakes

::: {.incremental style="font-size: 0.9em;"}
1. **Wrong projection**
   - Solution: Use appropriate projection (Albers, Mercator, etc.)

2. **Too many colors** in choropleth
   - Solution: Limit to 5-7 bins, use sequential scales

3. **Overlapping points** obscure patterns
   - Solution: Use transparency, jittering, or heat maps

4. **No context** (missing borders/labels)
   - Solution: Add state/country borders, major cities

5. **Color-blind unfriendly** palettes
   - Solution: Use viridis, ColorBrewer palettes

6. **Misleading scales**
   - Solution: Show scale bar, use consistent scales across facets
:::

## Geographic Data Sources

::: {style="font-size: 0.9em;"}
**Common Data Sources:**

::: {.incremental}
- **R packages:** `maps`, `rnaturalearth`, `tigris` (US census)
- **Shapefiles:** Government GIS data
- **APIs:** Google Maps, OpenStreetMap
- **Coordinates:** GPS, geocoding services

**Data Requirements:**

1. **Choropleth:** Region IDs matching map data
2. **Point maps:** Latitude and longitude
3. **Heat maps:** Many coordinate pairs
4. **All:** Consistent coordinate reference system (CRS)

**Tip:** Always verify coordinate system and projection!
:::
:::

## Practice Exercise

**Your Turn:**

1. Create a choropleth map showing mean BMI by Region
2. Build a point map of patients colored by SmokingStatus
3. Create a density heat map of patient locations
4. Make faceted maps showing patient distribution by Gender

**Bonus Challenge:**

Create an interactive leaflet map with popups showing patient details (ID, BP, BMI, Treatment).

## Summary

::: {style="font-size: 0.9em;"}
**What You've Learned:**

::: {.incremental}
1. **Choropleth maps** show values across regions using color
2. **Point/bubble maps** display individual locations and values
3. **Heat maps** reveal density and concentration patterns
4. **Faceted maps** enable categorical comparisons
5. **Interactive maps** enhance exploration

**Key Principle:**

> Geographic visualization reveals spatial patterns that tables and charts cannot show
:::
:::

## Next Steps {.center}

**You've completed Intermediate Level!**

**Next:** Advanced Level

- Categorical Advanced (specialized plots)
- Continuous Advanced (complex distributions)
- Time Series Advanced (forecasting, decomposition)
- And more!

You now have tools to create professional, publication-quality visualizations! 🎉
