---
title: "Continuous Plots - BASIC"
subtitle: "Visualizing Numeric Data"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Continuous Data {.center}

## What is Continuous Data?

::: {.columns}
::: {.column width="50%"}
**Definition:**

Data that represents **numeric measurements** or **quantities**

- Can take any value within a range
- Quantitative rather than qualitative
- Often derived from measurements
:::

::: {.column width="50%"}
**Examples in Our Dataset:**

- Age (18-79 years)
- BMI (11.4-37.3)
- Cholesterol levels
- Blood Sugar levels
- Blood Pressure (Systolic & Diastolic)
- Latitude & Longitude
:::
:::

## Analytical Goals for Continuous Data

::: {style="font-size: 0.9em;"}
**What questions can we answer?**

::: {.incremental}
- 🟡 **Distribution** - What's the spread? Shape? Center?
- ⚫ **Correlation** - How do two variables relate to each other?
:::

**Charts we'll learn:**

1. Histogram
2. Density Plot
3. Scatter Plot
4. Box Plot (covered in Category vs. Continuous)
:::

## Setup: Load Libraries and Data

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-3|5-6|8-11|13-15"

# Load required libraries
library(ggplot2)    # For creating plots
library(dplyr)      # For data manipulation

# Load the dataset
data <- readxl::read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Handle missing data: select complete cases for key numeric variables
data_clean <- data %>%
  filter(!is.na(BMI), !is.na(Cholesterol), !is.na(Age))  # Remove missing values

# Preview numeric variables
summary(data_clean %>% 
  select(Age, BMI, Cholesterol, BloodSugar, SysBP_0m, DiaBP_0m))
```

## Setup: Load Libraries and Data {.smaller}

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(readxl)

data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")
data_clean <- data %>% filter(!is.na(BMI), !is.na(Cholesterol), !is.na(Age))

summary(data_clean %>% select(Age, BMI, Cholesterol, BloodSugar, SysBP_0m, DiaBP_0m))
```

---

# Histogram

## Histogram - Code

🟡 **Goal:** Distribution (show frequency across bins)

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3|4-5|6|7-9"

# Create histogram showing age distribution
ggplot(data_clean, aes(x = Age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +  # 30 bins
  labs(title = "Distribution of Patient Ages",
       x = "Age (years)", y = "Frequency (Number of Patients)") +
  theme_minimal(base_family = "serif") +  # Serif font for professional look
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

::: {.notes}
- bins = 30 divides the data into 30 equal-width intervals
- Adjust bins to find optimal visualization (too few = loss of detail, too many = noise)
:::

## Histogram - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = Age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.8) +
  labs(title = "Distribution of Patient Ages",
       x = "Age (years)", y = "Frequency (Number of Patients)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

---

# Interpreting Histograms

## How to Read Histograms

**Using the previous plot:**

::: {.incremental style="font-size: 0.85em;"}
**OBSERVATION:**
- Distribution appears roughly uniform across age range
- Ages range from ~18 to ~79 years
- No clear peak (not normally distributed)
- Relatively even frequency across all age groups
- No obvious outliers or gaps

**INTERPRETATION:**
- Study successfully recruited across broad age spectrum
- No age group dominates the sample
- Good representation for age-stratified analysis
- Uniform distribution suggests deliberate balanced recruitment

**IMPLICATIONS:**
- Results generalizable across adult age range
- Can analyze age as continuous variable or create age groups
- Sufficient sample in all age ranges for subgroup analysis

**ACTIONS:**
- Consider age as covariate in treatment analysis
- May want to create age categories (e.g., 18-40, 41-60, 61+)
- Check if age interacts with treatment effect
:::
:::

---

# Histogram - With Statistics

## Histogram with Mean/Median Lines - Code

🟡 **Goal:** Distribution with central tendency markers

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-4|6-7|8-11|12-15|16-18"

# Calculate summary statistics
age_mean <- mean(data_clean$Age, na.rm = TRUE)  # Calculate mean
age_median <- median(data_clean$Age, na.rm = TRUE)  # Calculate median

# Create histogram with reference lines
ggplot(data_clean, aes(x = Age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(aes(xintercept = age_mean), color = "red", 
             linetype = "dashed", size = 1) +  # Add mean line
  geom_vline(aes(xintercept = age_median), color = "darkgreen", 
             linetype = "solid", size = 1) +  # Add median line
  annotate("text", x = age_mean + 5, y = 15, 
           label = paste("Mean =", round(age_mean, 1)), 
           color = "red", family = "serif", size = 4) +  # Label mean
  annotate("text", x = age_median - 5, y = 15, 
           label = paste("Median =", round(age_median, 1)), 
           color = "darkgreen", family = "serif", size = 4) +  # Label median
  labs(title = "Age Distribution with Mean and Median",
       x = "Age (years)", y = "Frequency") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))
```

## Histogram with Statistics - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

age_mean <- mean(data_clean$Age, na.rm = TRUE)
age_median <- median(data_clean$Age, na.rm = TRUE)

ggplot(data_clean, aes(x = Age)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white", alpha = 0.7) +
  geom_vline(aes(xintercept = age_mean), color = "red", 
             linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = age_median), color = "darkgreen", 
             linetype = "solid", size = 1) +
  annotate("text", x = age_mean + 5, y = 15, 
           label = paste("Mean =", round(age_mean, 1)), 
           color = "red", family = "serif", size = 4) +
  annotate("text", x = age_median - 5, y = 15, 
           label = paste("Median =", round(age_median, 1)), 
           color = "darkgreen", family = "serif", size = 4) +
  labs(title = "Age Distribution with Mean and Median",
       x = "Age (years)", y = "Frequency") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

---

# Density Plot

## Density Plot - Code

🟡 **Goal:** Distribution (smooth curve, easier to compare)

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3|4-5|6-8"

# Create smooth density plot for BMI
ggplot(data_clean, aes(x = BMI)) +
  geom_density(fill = "coral", alpha = 0.6, color = "darkred", size = 1) +  # Smooth density
  labs(title = "Density Distribution of BMI",
       x = "Body Mass Index (BMI)", y = "Density") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12))
```

::: {.notes}
Density plots smooth out histogram bins, showing underlying distribution shape
Area under curve = 1 (100%)
:::

## Density Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = BMI)) +
  geom_density(fill = "coral", alpha = 0.6, color = "darkred", size = 1) +
  labs(title = "Density Distribution of BMI",
       x = "Body Mass Index (BMI)", y = "Density") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

---

# Scatter Plot

## Scatter Plot - Code

⚫ **Goal:** Correlation between two continuous variables

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3|4-5|6-8"

# Create scatter plot showing BMI vs. Cholesterol relationship
ggplot(data_clean, aes(x = BMI, y = Cholesterol)) +
  geom_point(alpha = 0.6, color = "darkblue", size = 3) +  # Add semi-transparent points
  labs(title = "Relationship between BMI and Cholesterol",
       x = "Body Mass Index (BMI)", y = "Cholesterol (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12))
```

::: {.notes}
- alpha = 0.6 makes points semi-transparent to show overlapping
- Look for patterns: positive correlation, negative, or no relationship
:::

## Scatter Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = BMI, y = Cholesterol)) +
  geom_point(alpha = 0.6, color = "darkblue", size = 3) +
  labs(title = "Relationship between BMI and Cholesterol",
       x = "Body Mass Index (BMI)", y = "Cholesterol (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

---

# Interpreting Scatter Plots

## How to Read Scatter Plots

**Using the previous plot:**

::: {.incremental style="font-size: 0.85em;"}
**OBSERVATION:**
- Points scattered across the plot area
- No clear linear pattern visible
- BMI ranges from ~15 to ~37
- Cholesterol ranges from ~120 to ~290
- Some clustering but no strong trend

**INTERPRETATION:**
- Weak or no correlation between BMI and Cholesterol in this sample
- Relationship appears to be random/independent
- Both variables vary independently
- No obvious outliers or extreme cases

**IMPLICATIONS:**
- BMI alone may not predict cholesterol levels
- Other factors likely influence cholesterol (diet, genetics, medication)
- Cannot use BMI as sole predictor of cholesterol risk
- Need multivariate analysis for better understanding

**CRITICAL WARNING:**
- **Correlation ≠ Causation**
- Even if correlation existed, wouldn't prove causal relationship
- Always consider confounding variables
- Use domain knowledge alongside statistical patterns
:::
:::

---

# Scatter Plot - Colored by Group

## Scatter Plot with Categories - Code

⚫ **Goal:** Correlation + Compare groups

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3-4|5-7|8-10"

# Scatter plot colored by Treatment Group
ggplot(data_clean, aes(x = Age, y = SysBP_0m, color = TreatmentGroup)) +
  geom_point(alpha = 0.7, size = 3) +  # Color points by group
  scale_color_manual(values = c("Control" = "#E74C3C", 
                                 "Intervention" = "#3498DB"),
                     name = "Treatment Group") +  # Custom colors
  labs(title = "Baseline Systolic BP by Age and Treatment Group",
       x = "Age (years)", 
       y = "Systolic Blood Pressure (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        legend.position = "right")
```

## Scatter Plot with Categories - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = Age, y = SysBP_0m, color = TreatmentGroup)) +
  geom_point(alpha = 0.7, size = 3) +
  scale_color_manual(values = c("Control" = "#E74C3C", 
                                 "Intervention" = "#3498DB"),
                     name = "Treatment Group") +
  labs(title = "Baseline Systolic BP by Age and Treatment Group",
       x = "Age (years)", 
       y = "Systolic Blood Pressure (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Quick Reference

## Continuous Plots - Summary

| Chart Type | Goal | When to Use | Code Complexity |
|------------|------|-------------|-----------------|
| **Histogram** | 🟡 Distribution | Show frequency distribution, find shape | Simple (3-6 lines) |
| **Histogram + Stats** | 🟡 Distribution | Add mean/median for context | Medium (8-12 lines) |
| **Density Plot** | 🟡 Distribution | Smooth distribution, compare groups | Simple (3-6 lines) |
| **Scatter Plot** | ⚫ Correlation | Show relationship between 2 variables | Simple (3-6 lines) |
| **Scatter + Color** | ⚫ Correlation + Groups | Compare correlations across groups | Medium (6-8 lines) |

## Common Distribution Shapes

::: {.columns}
::: {.column width="50%"}
**Normal (Bell Curve)**
- Symmetric around mean
- Mean ≈ Median
- Example: Heights

**Right Skewed**
- Long tail to the right
- Mean > Median  
- Example: Income

**Left Skewed**
- Long tail to the left
- Mean < Median
- Example: Age at retirement
:::

::: {.column width="50%"}
**Bimodal**
- Two peaks
- Two distinct groups
- Example: Customer segments

**Uniform**
- Flat distribution
- All values equally likely
- Example: Random numbers

**Outliers**
- Extreme values
- Check for data errors
- May need special handling
:::
:::

## Key Takeaways

::: {style="font-size: 0.9em;"}
::: {.incremental}
**Remember:**

1. **Choose bins wisely** - test different values for histograms
2. **Density vs. Histogram** - density better for comparing groups
3. **Scatter plots reveal relationships** - but don't assume causation
4. **Alpha transparency** - helps see overlapping points
5. **Always handle missing data** - document your approach

**Common Mistakes:**

- Too many/few histogram bins
- Not handling outliers
- Assuming correlation = causation
- Ignoring data transformations when needed
- Poor color choices for scatter plot groups
:::
:::

## Practice Exercise

**Your Turn:**

1. Create a histogram of `Cholesterol` with 20 bins
2. Add mean and median lines
3. Create a scatter plot of `BloodSugar` vs. `SysBP_0m`
4. Color the scatter plot by `Gender`

**Bonus:** Create a density plot comparing BMI between smokers and non-smokers

## Next Steps {.center}

**You've mastered Continuous - Basic!**

**Next:** Category vs. Continuous - Basic

Learn to combine categorical and continuous data with:
- Box plots
- Grouped bar charts
- Violin plots (in Intermediate)
