---
title: "Category vs. Continuous - BASIC"
subtitle: "Comparing Numeric Data Across Groups"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Category vs. Continuous {.center}

## What is Category vs. Continuous Data?

::: {.columns}
::: {.column width="50%"}
**Definition:**

Combining **categorical variables** (groups) with **continuous variables** (numeric measurements)

- Compare numeric values across categories
- Show distributions within groups
- Analyze differences between segments
:::

::: {.column width="50%"}
**Examples in Our Dataset:**

- BMI by Gender
- Blood Pressure by Treatment Group
- Cholesterol by Region
- Age by Smoking Status
- Blood Sugar by Exercise Frequency
:::
:::

## Analytical Goals

::: {style="font-size: 0.9em;"}
**What questions can we answer?**

::: {.incremental}
- 🟡 **Distribution** - How do values spread within each category?
- 🟢 **Ranking** - Which group has highest/lowest values?
- 🔴 **Part of Whole** - How do groups contribute to total?
- ⚫ **Correlation** - Do patterns differ across groups?
:::

**Charts we'll learn:**

1. Box Plot
2. Grouped Bar Chart
3. Stacked Bar Chart
:::

## Setup: Load Data

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-3|5-6|8-11"

# Load required libraries
library(ggplot2)
library(dplyr)

# Load the dataset
data <- readxl::read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Handle missing data for complete case analysis
data_clean <- data %>%
  filter(!is.na(ExerciseFreq), !is.na(ChronicDisease))  # Remove rows with missing values
```

## Data Preview {.smaller}

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(readxl)

data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")
data_clean <- data %>% filter(!is.na(ExerciseFreq), !is.na(ChronicDisease))

# Show summary by group
data_clean %>%
  group_by(TreatmentGroup) %>%
  summarise(
    Count = n(),
    Mean_Age = round(mean(Age), 1),
    Mean_BMI = round(mean(BMI), 2),
    Mean_SysBP = round(mean(SysBP_0m), 1)
  )
```

---

# Box Plot

## Box Plot - Code

🟡 **Goal:** Compare distributions across groups

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3|4-6|7-9"

# Create box plot comparing BMI by Treatment Group
ggplot(data_clean, aes(x = TreatmentGroup, y = BMI, fill = TreatmentGroup)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.size = 3) +  # Add box plots
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB"),
                    name = "Treatment Group") +  # Custom colors
  labs(title = "BMI Distribution by Treatment Group",
       x = "Treatment Group", y = "Body Mass Index (BMI)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        legend.position = "none")  # Hide legend (redundant with x-axis)
```

::: {.notes}
Box plot shows: minimum, Q1 (25%), median (50%), Q3 (75%), maximum, and outliers
:::

## Box Plot - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = TreatmentGroup, y = BMI, fill = TreatmentGroup)) +
  geom_boxplot(alpha = 0.7, outlier.color = "red", outlier.size = 3) +
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB"),
                    name = "Treatment Group") +
  labs(title = "BMI Distribution by Treatment Group",
       x = "Treatment Group", y = "Body Mass Index (BMI)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")
```

---

# Interpreting Box Plots

## How to Read Box Plots

**Using the previous plot:**

::: {.incremental style="font-size: 0.8em;"}
**Reading the 5-Number Summary:**
- **Bottom whisker:** Minimum (~12-15)
- **Bottom of box (Q1):** 25th percentile (~22)
- **Line in box (Median):** 50th percentile (~25)
- **Top of box (Q3):** 75th percentile (~28)
- **Top whisker:** Maximum (~35)
- **Red dots:** Outliers beyond 1.5 × IQR

**OBSERVATION:**
- Both groups have similar median BMI (~25)
- Both groups show symmetric distributions
- Control group: slightly wider box (more variation)
- Few outliers in both groups
- Boxes largely overlap

**INTERPRETATION:**
- No substantial difference in BMI between treatment groups at baseline
- Random assignment appears successful (balanced groups)
- Similar BMI variability in both groups
- Treatment groups comparable on this key health metric

**IMPLICATIONS:**
- BMI unlikely to confound treatment effect
- Groups well-balanced for analysis
- Any post-treatment BMI differences likely due to treatment, not baseline

**ACTIONS:**
- Proceed with treatment comparison without BMI adjustment
- May still include BMI as covariate for precision
- Monitor for BMI changes during study
:::
:::

---

# Box Plot - Multiple Groups

## Box Plot with Multiple Categories - Code

🟡 **Goal:** Compare distributions across many groups

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|3-6|7-10|11-14"

# Compare Cholesterol across all regions
ggplot(data_clean, aes(x = reorder(Region, Cholesterol, FUN = median), 
                       y = Cholesterol, fill = Region)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +  # Box plots
  scale_fill_brewer(palette = "Set2", name = "Region") +  # Use color palette
  labs(title = "Cholesterol Levels by Region (Ordered by Median)",
       x = "Region", 
       y = "Cholesterol (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")  # X-axis provides category names
```

::: {.notes}
reorder() sorts regions by median cholesterol for easier comparison
:::

## Box Plot - Multiple Groups Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

ggplot(data_clean, aes(x = reorder(Region, Cholesterol, FUN = median), 
                       y = Cholesterol, fill = Region)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 2) +
  scale_fill_brewer(palette = "Set2", name = "Region") +
  labs(title = "Cholesterol Levels by Region (Ordered by Median)",
       x = "Region", 
       y = "Cholesterol (mg/dL)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.position = "none")
```

---

# Grouped Bar Chart

## Grouped Bar Chart - Code

🟢 **Goal:** Compare means across categories and groups

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-5|7-9|10-13|14-16"

# Calculate mean systolic BP by Region and Treatment Group
bp_summary <- data_clean %>%
  group_by(Region, TreatmentGroup) %>%
  summarise(Mean_SysBP = mean(SysBP_0m, na.rm = TRUE),
            .groups = "drop")  # Calculate group means

# Create grouped bar chart
ggplot(bp_summary, aes(x = Region, y = Mean_SysBP, fill = TreatmentGroup)) +
  geom_col(position = "dodge", alpha = 0.8) +  # Bars side-by-side
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB"),
                    name = "Treatment Group") +
  labs(title = "Mean Baseline Systolic BP by Region and Treatment Group",
       x = "Region", 
       y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.title = element_text(size = 12),
        legend.position = "right")
```

## Grouped Bar Chart - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false

bp_summary <- data_clean %>%
  group_by(Region, TreatmentGroup) %>%
  summarise(Mean_SysBP = mean(SysBP_0m, na.rm = TRUE),
            .groups = "drop")

ggplot(bp_summary, aes(x = Region, y = Mean_SysBP, fill = TreatmentGroup)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB"),
                    name = "Treatment Group") +
  labs(title = "Mean Baseline Systolic BP by Region and Treatment Group",
       x = "Region", 
       y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Stacked Bar Chart

## Stacked Bar Chart - Code

🔴 **Goal:** Part of Whole + Comparison

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-4|6-8|9-13|14-17"

# Count patients by Region and Gender
region_gender <- data_clean %>%
  count(Region, Gender) %>%
  filter(Gender != "Other")  # Remove small category for clarity

# Create stacked bar chart
ggplot(region_gender, aes(x = Region, y = n, fill = Gender)) +
  geom_col(position = "stack", alpha = 0.8) +  # Stack bars
  scale_fill_manual(values = c("Female" = "#E91E63", 
                                "Male" = "#2196F3"),
                    name = "Gender") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            color = "white", family = "serif", fontface = "bold", size = 4) +  # Add counts
  labs(title = "Patient Count by Region and Gender",
       x = "Region", 
       y = "Number of Patients") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        legend.position = "right")
```

## Stacked Bar Chart - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6

region_gender <- data_clean %>%
  count(Region, Gender) %>%
  filter(Gender != "Other")

ggplot(region_gender, aes(x = Region, y = n, fill = Gender)) +
  geom_col(position = "stack", alpha = 0.8) +
  scale_fill_manual(values = c("Female" = "#E91E63", 
                                "Male" = "#2196F3"),
                    name = "Gender") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            color = "white", family = "serif", fontface = "bold", size = 4) +
  labs(title = "Patient Count by Region and Gender",
       x = "Region", 
       y = "Number of Patients") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Quick Reference

## Category vs. Continuous - Summary

| Chart Type | Goal | When to Use | Code Complexity |
|------------|------|-------------|-----------------|
| **Box Plot** | 🟡 Distribution | Compare distributions across 2-6 groups | Simple (4-6 lines) |
| **Box Plot (Multiple)** | 🟡 Distribution | Compare many groups, show ranking | Medium (6-8 lines) |
| **Grouped Bar** | 🟢 Ranking/Comparison | Compare means across categories | Medium (8-12 lines) |
| **Stacked Bar** | 🔴 Part of Whole | Show composition and totals | Medium (8-12 lines) |

## Understanding Box Plots

::: {.columns style="font-size: 0.85em;"}
::: {.column width="50%"}
**The Box:**
- Bottom edge: Q1 (25th percentile)
- Middle line: Median (50th percentile)
- Top edge: Q3 (75th percentile)
- Box height: IQR (Q3 - Q1)

**The Whiskers:**
- Extend to 1.5 × IQR
- Show data range (excluding outliers)
:::

::: {.column width="50%"}
**Outliers:**
- Points beyond whiskers
- May indicate:
  - Data entry errors
  - Natural extreme values
  - Different population

**Comparing Boxes:**
- Non-overlapping: likely different
- Overlapping: may not be different
- Check median positions
- Compare IQR widths
:::
:::

## Key Takeaways

::: {style="font-size: 0.9em;"}
::: {.incremental}
**Remember:**

1. **Box plots show distribution** - not just means
2. **Order categories** - by median or logically
3. **Grouped vs. Stacked bars** - different purposes
4. **Always show units** - in axis labels
5. **Handle outliers thoughtfully** - investigate but don't automatically remove

**Common Mistakes:**

- Too many categories in one plot (>8)
- Not ordering categories logically
- Confusing grouped and stacked bars
- Ignoring outliers without investigation
- Comparing means without showing variation
:::
:::

## Practice Exercise

**Your Turn:**

1. Create a box plot of `Age` by `SmokingStatus`
2. Order the smoking statuses by median age
3. Create a grouped bar chart showing mean `Cholesterol` by `Region` and `Gender`
4. Add error bars (standard error) to the grouped bar chart

**Bonus:** Create a stacked bar showing `ExerciseFreq` composition by `Region`

## Next Steps {.center}

**You've mastered Category vs. Continuous - Basic!**

**Next:** Time Series - Basic

Learn to visualize data over time with:
- Line charts
- Multiple time series
- Trend analysis
