---
title: "Time Series - BASIC"
subtitle: "Visualizing Data Over Time"
format:
  revealjs:
    theme: serif
    css: ../../assets/slide_deck.css
    include-after-body: footer.html
    slide-number: true
    navigation-mode: linear
    code-copy: true
    code-fold: true
    chalkboard: true
    preview-links: true
    center: true
    zoom: true
---

# Time Series Data {.center}

## What is Time Series Data?

::: {.columns}
::: {.column width="50%"}
**Definition:**

Data points indexed by **time** or **ordered sequence**

- Observations at regular intervals
- Temporal ordering matters
- Shows change over time
- Can reveal trends and patterns
:::

::: {.column width="50%"}
**Examples in Our Dataset:**

- Blood Pressure at 0, 3, 6 months
- Patient measurements over time
- Baseline dates
- Treatment progression
- Longitudinal health outcomes
:::
:::

## Analytical Goals for Time Series

::: {style="font-size: 0.9em;"}
**What questions can we answer?**

::: {.incremental}
- 🔵 **Evolution** - How does it change over time?
- 🟡 **Distribution** - What's the pattern across time periods?
- 🟢 **Ranking** - Which time period is highest/lowest?
:::

**Charts we'll learn:**

1. Simple Line Chart
2. Multiple Line Chart (comparing groups)
3. Line Chart with Points
:::

## Setup: Load and Prepare Data

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-4|6-7|9-17|19-21"

# Load required libraries
library(ggplot2)
library(dplyr)
library(tidyr)      # For data reshaping

# Load the dataset
data <- readxl::read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

# Reshape data from wide to long format for time series plotting
# Convert columns SysBP_0m, SysBP_3m, SysBP_6m to rows
bp_long <- data %>%
  select(PatientID, TreatmentGroup, SysBP_0m, SysBP_3m, SysBP_6m) %>%
  pivot_longer(cols = starts_with("SysBP"),
               names_to = "TimePoint",
               values_to = "SysBP") %>%  # Convert to long format
  mutate(Month = case_when(TimePoint == "SysBP_0m" ~ 0,
                           TimePoint == "SysBP_3m" ~ 3,
                           TimePoint == "SysBP_6m" ~ 6))  # Extract month number

# Preview the transformed data
head(bp_long)
```

## Data Preview {.smaller}

```{r}
#| echo: false
#| message: false
#| warning: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(readxl)

data <- read_excel("../../data/PublicHealth_BP_Dataset.xlsx")

bp_long <- data %>%
  select(PatientID, TreatmentGroup, SysBP_0m, SysBP_3m, SysBP_6m) %>%
  pivot_longer(cols = starts_with("SysBP"),
               names_to = "TimePoint",
               values_to = "SysBP") %>%
  mutate(Month = case_when(TimePoint == "SysBP_0m" ~ 0,
                           TimePoint == "SysBP_3m" ~ 3,
                           TimePoint == "SysBP_6m" ~ 6))

head(bp_long, 12)
```

---

# Simple Line Chart

## Simple Line Chart - Code

🔵 **Goal:** Show evolution over time (aggregate)

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-4|6-8|9-11|12-15"

# Calculate mean systolic BP at each time point
mean_bp <- bp_long %>%
  group_by(Month) %>%
  summarise(Mean_SysBP = mean(SysBP, na.rm = TRUE))  # Average across all patients

# Create line chart showing trend over time
ggplot(mean_bp, aes(x = Month, y = Mean_SysBP)) +
  geom_line(color = "steelblue", size = 1.5) +  # Draw line
  geom_point(color = "steelblue", size = 4) +  # Add points at each time point
  scale_x_continuous(breaks = c(0, 3, 6)) +  # Show all time points on x-axis
  labs(title = "Mean Systolic Blood Pressure Over 6 Months",
       x = "Month", 
       y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

## Simple Line Chart - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false

mean_bp <- bp_long %>%
  group_by(Month) %>%
  summarise(Mean_SysBP = mean(SysBP, na.rm = TRUE))

ggplot(mean_bp, aes(x = Month, y = Mean_SysBP)) +
  geom_line(color = "steelblue", size = 1.5) +
  geom_point(color = "steelblue", size = 4) +
  scale_x_continuous(breaks = c(0, 3, 6)) +
  labs(title = "Mean Systolic Blood Pressure Over 6 Months",
       x = "Month", 
       y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10))
```

---

# Interpreting Line Charts

## How to Read Line Charts

**Using the previous plot:**

::: {.incremental style="font-size: 0.85em;"}
**OBSERVATION:**
- Baseline (Month 0): Mean SysBP ~135 mmHg
- Month 3: Mean SysBP ~129 mmHg (decrease of ~6 mmHg)
- Month 6: Mean SysBP ~130 mmHg (slight increase from Month 3)
- Overall trend: Downward from baseline
- Small rebound between months 3 and 6

**INTERPRETATION:**
- Blood pressure decreased during first 3 months of study
- Effect stabilized or slightly reversed by month 6
- Overall improvement of ~5 mmHg from baseline
- Intervention may have time-dependent effects

**IMPLICATIONS:**
- Treatment appears effective in short term (3 months)
- Longer-term effects need investigation
- 5 mmHg reduction clinically meaningful
- May need to compare Control vs. Intervention separately

**ACTIONS:**
- Separate analysis by treatment group needed
- Consider reasons for Month 6 rebound
- Evaluate if maintenance strategies needed
- Assess clinical significance (5 mmHg reduction = ~10% stroke risk ↓)
:::
:::

---

# Multiple Line Chart

## Multiple Line Chart - Code

🔵 **Goal:** Compare evolution across groups

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-4|6-9|10-13|14-16"

# Calculate mean BP by treatment group and time
mean_bp_group <- bp_long %>%
  group_by(Month, TreatmentGroup) %>%
  summarise(Mean_SysBP = mean(SysBP, na.rm = TRUE), .groups = "drop")

# Create line chart with separate lines for each group
ggplot(mean_bp_group, aes(x = Month, y = Mean_SysBP, 
                          color = TreatmentGroup, group = TreatmentGroup)) +
  geom_line(size = 1.5) +  # Draw lines
  geom_point(size = 4) +  # Add points
  scale_color_manual(values = c("Control" = "#E74C3C", 
                                 "Intervention" = "#3498DB"),
                     name = "Treatment Group") +  # Custom colors
  scale_x_continuous(breaks = c(0, 3, 6)) +
  labs(title = "Systolic BP by Treatment Group Over Time",
       x = "Month", y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        legend.position = "right")
```

## Multiple Line Chart - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false

mean_bp_group <- bp_long %>%
  group_by(Month, TreatmentGroup) %>%
  summarise(Mean_SysBP = mean(SysBP, na.rm = TRUE), .groups = "drop")

ggplot(mean_bp_group, aes(x = Month, y = Mean_SysBP, 
                          color = TreatmentGroup, group = TreatmentGroup)) +
  geom_line(size = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Control" = "#E74C3C", 
                                 "Intervention" = "#3498DB"),
                     name = "Treatment Group") +
  scale_x_continuous(breaks = c(0, 3, 6)) +
  labs(title = "Systolic BP by Treatment Group Over Time",
       x = "Month", y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Line Chart with Confidence Intervals

## Line Chart with Error Bars - Code

🔵 **Goal:** Show trend with uncertainty

```{r}
#| echo: true
#| eval: false
#| code-line-numbers: "|1-7|9-13|14-17|18-21"

# Calculate mean and standard error by group
bp_summary <- bp_long %>%
  group_by(Month, TreatmentGroup) %>%
  summarise(Mean_SysBP = mean(SysBP, na.rm = TRUE),
            SE = sd(SysBP, na.rm = TRUE) / sqrt(n()),  # Standard error
            .groups = "drop")

# Create line chart with error ribbons
ggplot(bp_summary, aes(x = Month, y = Mean_SysBP, 
                       color = TreatmentGroup, fill = TreatmentGroup)) +
  geom_ribbon(aes(ymin = Mean_SysBP - SE, ymax = Mean_SysBP + SE),
              alpha = 0.2, color = NA) +  # Add confidence bands
  geom_line(size = 1.5) +  # Draw lines on top
  geom_point(size = 4) +  # Add points
  scale_color_manual(values = c("Control" = "#E74C3C", 
                                 "Intervention" = "#3498DB")) +
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB")) +
  scale_x_continuous(breaks = c(0, 3, 6)) +
  labs(title = "Systolic BP Over Time with Standard Error",
       x = "Month", y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        legend.position = "right")
```

## Line Chart with Error Bars - Plot

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 6
#| message: false

bp_summary <- bp_long %>%
  group_by(Month, TreatmentGroup) %>%
  summarise(Mean_SysBP = mean(SysBP, na.rm = TRUE),
            SE = sd(SysBP, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

ggplot(bp_summary, aes(x = Month, y = Mean_SysBP, 
                       color = TreatmentGroup, fill = TreatmentGroup)) +
  geom_ribbon(aes(ymin = Mean_SysBP - SE, ymax = Mean_SysBP + SE),
              alpha = 0.2, color = NA) +
  geom_line(size = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Control" = "#E74C3C", 
                                 "Intervention" = "#3498DB"),
                     name = "Treatment\nGroup") +
  scale_fill_manual(values = c("Control" = "#E74C3C", 
                                "Intervention" = "#3498DB"),
                    name = "Treatment\nGroup") +
  scale_x_continuous(breaks = c(0, 3, 6)) +
  labs(title = "Systolic BP Over Time with Standard Error",
       x = "Month", y = "Mean Systolic BP (mmHg)") +
  theme_minimal(base_family = "serif") +
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
        axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 11, face = "bold"),
        legend.text = element_text(size = 10))
```

---

# Quick Reference

## Time Series - Summary

| Chart Type | Goal | When to Use | Code Complexity |
|------------|------|-------------|-----------------|
| **Simple Line** | 🔵 Evolution | Single metric over time | Simple (4-8 lines) |
| **Multiple Lines** | 🔵 Evolution + Compare | Compare groups over time | Medium (6-10 lines) |
| **Line + Confidence** | 🔵 Evolution + Uncertainty | Show trends with error | Medium (10-15 lines) |

## Key Time Series Concepts

::: {.columns style="font-size: 0.85em;"}
::: {.column width="50%"}
**Trend:**
- Overall direction (up, down, flat)
- Long-term pattern
- Calculate: moving average, regression

**Seasonality:**
- Regular periodic fluctuations
- Daily, weekly, monthly patterns
- Example: Higher BP in winter

**Volatility:**
- Amount of variation
- High = unstable, Low = stable
- Measure: standard deviation
:::

::: {.column width="50%"}
**Data Preparation:**
- Wide format: One column per time point
- Long format: One row per observation
- Use `pivot_longer()` to convert

**Best Practices:**
- Always show all time points
- Use consistent time intervals
- Label axes clearly with units
- Consider adding reference lines
- Show confidence intervals when relevant
:::
:::

## Key Takeaways

::: {style="font-size: 0.9em;"}
::: {.incremental}
**Remember:**

1. **Connect the dots** - use geom_line() to show continuity
2. **Add points** - helps identify exact time points
3. **Multiple lines need legends** - make groups clear
4. **Show uncertainty** - confidence intervals add credibility
5. **Wide to long format** - essential for time series plotting

**Common Mistakes:**

- Not connecting points (scatter instead of line)
- Too many lines (>5 becomes cluttered)
- Inconsistent time intervals without noting
- Missing baseline comparisons
- Not handling missing time points
:::
:::

## Practice Exercise

**Your Turn:**

1. Create a line chart of mean `DiaBP` (diastolic BP) over 0, 3, 6 months
2. Add separate lines for Control vs. Intervention
3. Include confidence intervals (± SE)
4. Add a horizontal reference line at 80 mmHg (hypertension threshold)

**Bonus:** Create a line chart showing mean `BMI` changes over time by `Gender`

## Next Steps {.center}

**You've completed all BASIC level decks!**

🎉 **Congratulations!** 

**Next:** Intermediate Level

Move to Module 2 for enhanced visualizations:
- Categorical Intermediate (Lollipop, Mosaic plots)
- Continuous Intermediate (Violin, Ridge plots)
- Advanced time series
- Geographic maps
- Network diagrams
